---
title: "Building the Analytic Spine — Navrongo HDSS Extract"
subtitle: "Heat Stress and Perinatal Outcomes in Ghana"
author: "Vincent"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  pdf_document:
    toc: yes
    toc_depth: 3
    number_sections: yes
    keep_tex: no
    latex_engine: xelatex
  html_document:
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: false
      smooth_scroll: true
    number_sections: yes
    code_folding: show
    df_print: kable
    theme: flatly
    highlight: tango
    fig_caption: yes
  github_document:
    toc: yes
    toc_depth: 3
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  message    = FALSE,
  warning    = FALSE,
  comment    = "#>",
  fig.align  = "center",
  fig.width  = 7,
  fig.height = 4.5,
  dpi        = 110
)
options(scipen = 999)
```

# Purpose and scope

This document implements steps (ii)–(iv) of the analytic plan accompanying the
proposal *Heat Stress and Perinatal Outcomes in Ghana: A Spatio-Temporal
Analysis* (Version 2.0). It produces the cleaned analytic spine that all
downstream modelling scripts will consume; it does not itself fit any
statistical model.

The three operations executed here are:

- **(ii) Deduplication.** The raw extract contains one row per maternal
  residency episode, which yields multiple rows per child when a mother's
  residency history spans more than one episode. The spine is reduced to one
  row per unique child, and the maternal residency episodes are retained in a
  separate long-format table that downstream survival or migration-history
  models can join back to.
- **(iii) Recoding of HDSS missing-data sentinels.** Two integer sentinels are
  in use across the HDSS extract — `88` and `99` — per the data-management
  conventions confirmed by NHRC. Both are recoded to `NA` for `ceb`,
  `prgmonth`, and `numanc`. The value `0` in `child_bwgt` is treated as
  "not weighed" rather than as a real observation. Biologically implausible
  birth weights (>6 kg or <0.5 kg) are also set to `NA`, as are blank cells
  in `mother_mstatus`.
- **(iv) Construction of derived analytic variables.** Mortality indicators
  (day-0, early-neonatal, neonatal, infant, under-five), birth-weight
  categories (LBW, VLBW), maternal-age and parity categories, season of
  delivery on both the three-season and two-season conventions for northern
  Ghana, reconstructed pregnancy windows for downstream DLNM linkage, and a
  flag for the **post-2012 ANC-reliable sub-cohort** used in the
  care-modifier sub-analyses.

**Analytic window.** Following the workflow agreed with NHRC, the primary
analytic window covers *all available birth-outcome records* in the full
extract, rather than a fixed ten-year span. This decision reflects the
relative scarcity of perinatal-mortality events and the value of preserving
the full longitudinal record for the main outcome models. The antenatal-care
variables (`numanc`, `prgmonth`, `ancatt`) are reliably populated only from
2013 onward; analyses that adjust for or stratify on ANC are therefore
restricted to a 2013-onward sub-cohort, with a binary flag (`anc_reliable`)
constructed in the spine for that purpose.

The numeric category codes for `assist_del`, `child_birth_place`,
`delivery`, `cutcod`, `cordres1`, `seventtype`, and `eeventtype` are kept as
integers pending the NHRC value-label codebook, which will be applied in a
follow-up labelling chunk once received from the Centre.

# Setup

```{r packages}
# Install any missing packages on first run:
# install.packages(c("haven","dplyr","tidyr","lubridate","stringr",
#                    "readr","knitr","kableExtra"))
library(haven)        # read_dta
library(dplyr)
library(tidyr)
library(lubridate)
library(stringr)
library(readr)
library(knitr)
library(kableExtra)

# Adjust these paths to match the project layout.
IN_PATH <- "vincent_phd_data_30042026.dta"
OUT_DIR <- "."
dir.create(OUT_DIR, showWarnings = FALSE, recursive = TRUE)
```

A small helper used throughout the document to render diagnostic tables in a
consistent style:

```{r helper}
nice_table <- function(x, caption = NULL, full_width = FALSE) {
  kable(x, caption = caption, format.args = list(big.mark = ",")) |>
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                  full_width = full_width, position = "left")
}
```

# Load raw data

`haven::read_dta()` preserves the Stata variable labels and value labels as
attributes, so they remain accessible for documentation while we operate on
the underlying numeric codes.

```{r load-raw}
raw <- read_dta(IN_PATH)

tibble(metric = c("rows", "columns",
                  "unique mothers (motherid)",
                  "unique children (child_id)",
                  "unique compounds (locationid)"),
       value  = c(nrow(raw), ncol(raw),
                  dplyr::n_distinct(raw$motherid),
                  dplyr::n_distinct(raw$child_id),
                  dplyr::n_distinct(raw$locationid))) |>
  nice_table(caption = "Raw extract — basic dimensions")
```

The difference between rows (15,639) and unique children (14,406) reflects
the residency-episode duplication that step (ii) will resolve.

# Step (iii): Recode HDSS missing-data sentinels

Per the NHRC data-management conventions, the integers `88` and `99` are
both used as missing-data sentinels in the parity (`ceb`) and antenatal-care
(`prgmonth`, `numanc`) fields. The value `0` in `child_bwgt` denotes a child
who was not weighed rather than a child who weighed zero kilograms. A small
number of birth weights also fall outside biologically plausible bounds and
are treated as missing.

```{r recode}
df <- raw |>
  mutate(
    ceb            = if_else(ceb      %in% c(88, 99), NA_real_, ceb),
    prgmonth       = if_else(prgmonth %in% c(88, 99), NA_real_, prgmonth),
    numanc         = if_else(numanc   %in% c(88, 99), NA_real_, numanc),
    child_bwgt     = if_else(child_bwgt == 0 | child_bwgt < 0.5 |
                               child_bwgt > 6.0,
                             NA_real_, child_bwgt),
    mother_mstatus = if_else(str_trim(mother_mstatus) == "",
                             NA_character_, mother_mstatus)
  )
```

Effect of the recode on the non-missing counts of the affected fields:

```{r recode-summary}
tibble(
  variable     = c("ceb", "prgmonth", "numanc", "child_bwgt", "mother_mstatus"),
  before_n     = c(sum(!is.na(raw$ceb)),
                   sum(!is.na(raw$prgmonth)),
                   sum(!is.na(raw$numanc)),
                   sum(!is.na(raw$child_bwgt)),
                   sum(raw$mother_mstatus != "" & !is.na(raw$mother_mstatus))),
  after_n      = c(sum(!is.na(df$ceb)),
                   sum(!is.na(df$prgmonth)),
                   sum(!is.na(df$numanc)),
                   sum(!is.na(df$child_bwgt)),
                   sum(!is.na(df$mother_mstatus))),
  n_recoded_NA = before_n - after_n
) |>
  nice_table(caption = "Non-missing counts before and after sentinel recoding")
```

# Step (ii.a): Build the residency-episode table

Each maternal residency episode is kept on its own row in a separate
long-format table. This file is the join key for any analysis that needs
in-migration or out-migration timing.

```{r residency-episodes}
residency_episodes <- df |>
  distinct(motherid, sdate, edate, seventtype, eeventtype,
           locationid, location) |>
  arrange(motherid, sdate) |>
  mutate(episode_id = row_number(),
         sdate = as.Date(sdate),
         edate = as.Date(edate))

tibble(metric = c("episode rows", "unique mothers represented"),
       value  = c(nrow(residency_episodes),
                  dplyr::n_distinct(residency_episodes$motherid))) |>
  nice_table(caption = "Residency-episode table — dimensions")
```

A small preview of the table (sorted by mother and episode start date), to
confirm that the structure is as expected:

```{r residency-preview}
head(residency_episodes, 6) |>
  nice_table(caption = "First six rows of the residency-episode table")
```

# Step (ii.b): Deduplicate to one row per child

Where multiple rows share a `child_id`, the surviving row is chosen by two
ordered criteria:

1. Lowest count of missing values across the analytic covariate set.
2. Tie-break on the most recent `sdate` (the latest residency context is
   closest in time to the heat-exposure window of interest).

```{r dedup}
analytic_vars <- c("mother_educ", "ses_index", "mother_mstatus",
                   "child_bwgt", "child_birth_place", "delivery",
                   "ancatt", "prgmonth", "numanc",
                   "first_nb", "assist_del", "cutcod", "cordres1")

spine_pre <- df |>
  mutate(.miss_score = rowSums(is.na(across(all_of(analytic_vars))))) |>
  arrange(child_id, .miss_score, desc(sdate)) |>
  distinct(child_id, .keep_all = TRUE) |>
  dplyr::select(-.miss_score)

stopifnot(!anyDuplicated(spine_pre$child_id))

tibble(metric = c("rows in spine (= unique children)",
                  "rows in raw extract",
                  "rows removed by deduplication"),
       value  = c(nrow(spine_pre),
                  nrow(df),
                  nrow(df) - nrow(spine_pre))) |>
  nice_table(caption = "Effect of child-level deduplication")
```

# Step (iv): Derived analytic variables

Several conventions used in this block deserve explicit mention.

- Age at death is computed as integer days from `child_dob` to `child_dod`.
  Children without a recorded death date are coded as `NA` for age at death,
  and as `0` (not `NA`) for the binary mortality indicators, so that those
  indicators are NA-safe for downstream regression.
- The 280-day term-gestation assumption used to back-date pregnancy start is
  a working default. Sensitivity analyses should also fit results at 259 and
  294 days to bracket gestational-age error.
- Northern Ghana climatology follows the conventional three-season split:
  *dry-cool* (Nov–Mar), *hot-dry* (Apr–May), and *wet* (Jun–Oct). A
  two-level wet-vs-dry version is also retained for parsimonious models.
- Parity is taken from `ceb` where available. Because `ceb` is missing for
  approximately 52% of records, a binary nulliparous proxy is also computed
  from the fully-populated `first_nb` field.

```{r derive}
spine <- spine_pre |>
  mutate(
    # ---- Dates ----
    mother_dob = as.Date(mother_dob),
    child_dob  = as.Date(child_dob),
    child_dod  = as.Date(child_dod),
    sdate      = as.Date(sdate),
    edate      = as.Date(edate),

    # ---- Age at death (days) ----
    age_death_days = as.integer(child_dod - child_dob),

    # ---- Mortality indicators (NA-safe) ----
    died_day0     = if_else(!is.na(age_death_days) & age_death_days == 0,         1L, 0L),
    died_enn      = if_else(!is.na(age_death_days) & age_death_days >= 0
                            & age_death_days <= 7,                                1L, 0L),
    died_neonatal = if_else(!is.na(age_death_days) & age_death_days >= 0
                            & age_death_days <= 28,                               1L, 0L),
    died_infant   = if_else(!is.na(age_death_days) & age_death_days >= 0
                            & age_death_days <= 365,                              1L, 0L),
    died_u5       = if_else(!is.na(age_death_days) & age_death_days >= 0
                            & age_death_days <= 1826,                             1L, 0L),

    # ---- Birth-weight derivatives ----
    lbw  = if_else(!is.na(child_bwgt), as.integer(child_bwgt <  2.5), NA_integer_),
    vlbw = if_else(!is.na(child_bwgt), as.integer(child_bwgt <  1.5), NA_integer_),

    # ---- Temporal markers ----
    birth_year  = year(child_dob),
    birth_month = month(child_dob),
    season3 = case_when(
      birth_month %in% c(11, 12, 1, 2, 3) ~ "dry_cool",
      birth_month %in% c(4, 5)            ~ "hot_dry",
      birth_month %in% c(6, 7, 8, 9, 10)  ~ "wet",
      TRUE                                ~ NA_character_),
    season2 = if_else(birth_month %in% 6:10, "wet", "dry"),

    # ---- Maternal age and parity ----
    mat_age_yrs = as.numeric(child_dob - mother_dob) / 365.25,
    mat_age_cat = cut(mat_age_yrs,
                      breaks = c(-Inf, 19.999, 24, 29, 34, Inf),
                      labels = c("<20", "20-24", "25-29", "30-34", "35+"),
                      right  = TRUE),
    parity_cat = cut(ceb,
                     breaks = c(0, 1, 2, 4, Inf),
                     labels = c("nulliparous", "1", "2-4", "5+"),
                     include.lowest = TRUE, right = TRUE),
    nulliparous_proxy = if_else(first_nb == 1, 1L, 0L),

    # ---- Education collapsed ----
    educ_any_secondary = if_else(!is.na(mother_educ),
                                 as.integer(mother_educ >= 2),
                                 NA_integer_),

    # ---- Reconstructed pregnancy windows (280-day term gestation) ----
    preg_start_assumed = child_dob - days(280),
    preg_t3_start      = child_dob - days(98),
    preg_last30        = child_dob - days(30),
    preg_last7         = child_dob - days(7),

    # ---- ANC-reliable sub-cohort flag ----
    # The antenatal-care fields are populated reliably only for births from
    # 2013 onward; analyses that adjust for or stratify on ANC are restricted
    # to this sub-cohort. The PRIMARY analytic window uses all available
    # birth-outcome records (no upper or lower year restriction).
    anc_reliable = as.integer(birth_year >= 2013),

    # ---- Labelled birth outcome ----
    outcome_class = case_when(
      child_birth_type == 1 ~ "abortion",
      child_birth_type == 2 ~ "miscarriage",
      child_birth_type == 3 ~ "live_birth",
      child_birth_type == 4 ~ "stillbirth",
      child_birth_type == 5 ~ "neonatal_death_at_record",
      child_birth_type == 6 ~ "censored_pregnancy",
      child_birth_type == 7 ~ "ongoing_pregnancy",
      TRUE                  ~ NA_character_)
  )
```

# Diagnostics

## Mortality event counts

The full-file event counts, before any analytic-window restriction:

```{r mortality-full}
spine |>
  summarise(
    across(c(died_day0, died_enn, died_neonatal, died_infant, died_u5), sum)
  ) |>
  pivot_longer(everything(), names_to = "indicator", values_to = "events") |>
  mutate(definition = recode(indicator,
    died_day0     = "Day-0 death (death on day of birth)",
    died_enn      = "Early-neonatal death (0–7 days)",
    died_neonatal = "Neonatal death (0–28 days)",
    died_infant   = "Infant death (0–365 days)",
    died_u5       = "Under-five death (0–1,826 days)"
  )) |>
  dplyr::select(indicator, definition, events) |>
  nice_table(caption = "Mortality event counts, full file (1984–2023)")
```

## ANC-reliable sub-cohort (2013 onward)

The 2013-onward sub-cohort is the operative one for any analysis that requires
the ANC variables, since these fields are systematically blank for births
before 2012. The full analytic window — used for the primary perinatal
outcome models — places no upper or lower year restriction beyond what is
present in the full extract delivered by NHRC.

```{r sub-cohort}
sub <- spine |> filter(anc_reliable == 1)

tibble(
  metric = c("n children in ANC-reliable sub-cohort",
             "early-neonatal deaths (0–7 days)",
             "neonatal deaths (0–28 days)",
             "infant deaths (0–365 days)",
             "records with valid birth weight",
             "low birth weight (<2.5 kg)",
             "records with non-missing numanc",
             "records with valid parity (ceb)"),
  value  = c(nrow(sub),
             sum(sub$died_enn,       na.rm = TRUE),
             sum(sub$died_neonatal,  na.rm = TRUE),
             sum(sub$died_infant,    na.rm = TRUE),
             sum(!is.na(sub$child_bwgt)),
             sum(sub$lbw,            na.rm = TRUE),
             sum(!is.na(sub$numanc)),
             sum(!is.na(sub$ceb)))
) |>
  nice_table(caption = "Event and observation counts in the ANC-reliable (2013-onward) sub-cohort")
```

## Distribution checks

Season of delivery in the ANC-reliable (2013-onward) sub-cohort, on both the
three-season and the two-season conventions:

```{r season-dist}
sub |>
  count(season3, name = "n") |>
  arrange(desc(n)) |>
  nice_table(caption = "Season of delivery (three-season), ANC-reliable sub-cohort")

sub |>
  count(season2, name = "n") |>
  arrange(desc(n)) |>
  nice_table(caption = "Season of delivery (wet/dry), ANC-reliable sub-cohort")
```

Maternal age category, full file (a useful background distribution to keep in
mind when interpreting age-stratified outputs later):

```{r mat-age-dist}
spine |>
  count(mat_age_cat, name = "n") |>
  arrange(mat_age_cat) |>
  nice_table(caption = "Maternal age category, full file")
```

# Spine codebook

The variable list below documents the derived analytic spine. It is also
saved to `spine_codebook.csv` for reference.

```{r codebook}
spine_codebook <- tribble(
  ~variable,             ~description,
  "child_id",            "Unique child identifier (NHRC).",
  "motherid",            "Unique mother identifier.",
  "locationid",          "Compound identifier (8,472 unique in this extract).",
  "location",            "Rural/Urban indicator (R/U).",
  "child_dob",           "Date of birth (delivery).",
  "child_dod",           "Date of death; NA if alive at last record.",
  "age_death_days",      "Integer days from child_dob to child_dod; NA if alive.",
  "died_day0",           "1 if death on day of birth (perinatal proxy).",
  "died_enn",            "1 if death within 0–7 days of birth (early neonatal).",
  "died_neonatal",       "1 if death within 0–28 days (neonatal).",
  "died_infant",         "1 if death within 0–365 days (infant).",
  "died_u5",             "1 if death within 0–1,826 days (under-five).",
  "gender",              "Sex of child (F/M).",
  "child_bwgt",          "Birth weight in kg; 0-coded values and outliers set NA.",
  "lbw",                 "1 if birth weight < 2.5 kg; NA if weight missing.",
  "vlbw",                "1 if birth weight < 1.5 kg; NA if weight missing.",
  "child_birth_place",   "Place of birth (1–4 codes; codebook needed).",
  "delivery",            "Mode of delivery (1–5 codes; codebook needed).",
  "assist_del",          "Delivery-assistance code (1–6; codebook needed).",
  "cutcod",              "Material used to cut umbilical cord.",
  "cordres1",            "Material used to dress umbilical stump.",
  "ancatt",              "ANC attendance (1=yes, 2=no); reliable only 2013+.",
  "prgmonth",            "Pregnancy month at first ANC visit; reliable only 2013+.",
  "numanc",              "Number of ANC visits during pregnancy; reliable only 2013+.",
  "first_nb",            "1 if first live birth of mother; 2 otherwise.",
  "ceb",                 "Children ever born; 88 and 99 are HDSS missing sentinels (recoded NA).",
  "parity_cat",          "Parity categories from ceb; NA where ceb missing.",
  "nulliparous_proxy",   "1 if first_nb==1 (use when ceb missing).",
  "mother_dob",          "Mother date of birth.",
  "mat_age_yrs",         "Maternal age at delivery (years).",
  "mat_age_cat",         "Maternal-age categories: <20, 20-24, 25-29, 30-34, 35+.",
  "mother_educ",         "Maternal education (0=none,1=primary,2=JHS,3=SHS,4=tertiary).",
  "educ_any_secondary",  "1 if mother_educ >= 2 (JHS or higher).",
  "mother_mstatus",      "Marital status (MAR, DIV, WID, DTH; blanks recoded NA).",
  "ses_index",           "SES quintile (1=poorest – 5=least poor).",
  "birth_year",          "Calendar year of birth.",
  "birth_month",         "Calendar month of birth.",
  "season3",             "Season of delivery (dry_cool / hot_dry / wet).",
  "season2",             "Season of delivery (dry / wet).",
  "preg_start_assumed",  "Assumed pregnancy start (child_dob - 280 days).",
  "preg_t3_start",       "Start of third trimester (child_dob - 98 days).",
  "preg_last30",         "30 days before delivery.",
  "preg_last7",          "7 days before delivery.",
  "anc_reliable",        "1 if birth in 2013 onward (ANC-reliable sub-cohort).",
  "outcome_class",       "Labelled child_birth_type."
)

nice_table(spine_codebook, caption = "Spine codebook")
```

# Persist outputs

The spine and the residency-episode table are written in both R-native
(`.rds`) and text (`.csv`) formats so that downstream scripts in R, Stata,
Python, or any other environment can read them without modification.

```{r save, results='hold'}
saveRDS(spine,              file.path(OUT_DIR, "spine.rds"))
saveRDS(residency_episodes, file.path(OUT_DIR, "residency_episodes.rds"))
write_csv(spine,              file.path(OUT_DIR, "spine.csv"))
write_csv(residency_episodes, file.path(OUT_DIR, "residency_episodes.csv"))
write_csv(spine_codebook,     file.path(OUT_DIR, "spine_codebook.csv"))

cat("Files written to ", normalizePath(OUT_DIR), ":\n  ",
    paste(c("spine.rds", "residency_episodes.rds",
            "spine.csv", "residency_episodes.csv",
            "spine_codebook.csv"),
          collapse = "\n  "), sep = "")
```

# Expected diagnostic counts (sample dataset)

The script has been validated against the sample extract delivered on 30
April 2026. A successful run on that sample should reproduce the following
counts; any discrepancy indicates that something has changed in the input
file and should be investigated before downstream modelling proceeds. The
full-data run, performed by the Centre's analysts under the agreed
workflow, will produce different (larger) figures and an extended
birth-year range; the expected-counts table is preserved as a sample-stage
validation reference, not as a target for the full data.

```{r expected, echo=FALSE}
tibble(
  metric = c("rows in raw extract",
             "unique children",
             "unique mothers",
             "unique compounds",
             "birth-year range (sample)",
             "day-0 deaths (full file)",
             "early-neonatal deaths (full file)",
             "neonatal deaths (full file)",
             "infant deaths (full file)",
             "under-five deaths (full file)",
             "n children in ANC-reliable sub-cohort",
             "early-neonatal deaths, sub-cohort",
             "neonatal deaths, sub-cohort",
             "valid birth weights, sub-cohort",
             "LBW (<2.5 kg), sub-cohort"),
  expected_value = c("15,639", "14,406", "13,284", "8,472", "1984–2023",
                     "67", "166", "258", "636", "1,027",
                     "5,560", "32", "41", "3,190", "387")
) |>
  nice_table(caption = "Expected diagnostic counts for the SAMPLE dataset. Full-data figures will differ.")
```

# Session information

```{r session}
sessionInfo()
```

# Next steps

With the spine in hand, the immediate items are, in order:

1. **Apply the NHRC value-label codebook** for `assist_del`, `child_birth_place`,
   `delivery`, `cutcod`, `cordres1`, `seventtype`, and `eeventtype` once it is
   received from the Centre, and add a labelling chunk to this document that
   promotes these integer fields to labelled factors. NHRC has confirmed
   labels will be supplied where applicable.
2. Write `02_link_era5_utci.Rmd`, which (i) reads `spine.rds`,
   (ii) consumes the household-level geocodes that NHRC will release with
   the full extract, (iii) links each pregnancy to a daily UTCI series
   from ERA5-Land at the household coordinate, and (iv) writes the
   pregnancy-level exposure matrix to `exposure.rds`. Per the workflow
   agreed with NHRC, all climate-variable generation — ERA5-Land download,
   UTCI computation following Di Napoli et al. (2018), and gestational-window
   aggregation — is conducted within this script and executed by the
   Centre's analysts against the full extract.
3. Write `03_phase1_descriptive.Rmd`, `04_phase2_misclassification.Rmd`,
   `05_phase3_dlnm_perinatal.Rmd`, `06_phase4_casecrossover.Rmd`,
   `07_phase5_dlnm_birthweight.Rmd`, `08_phase6_effect_modification.Rmd`,
   and `09_phase7_sensitivity.Rmd`, in line with the Chapter 3 analytic plan.
   The effect-modification and ANC-related analyses use the 2013-onward
   `anc_reliable` sub-cohort; the primary perinatal-outcome models use the
   full birth-year range available in the NHRC extract.
