---
title: "Building the Analytic Spine — Navrongo HDSS Extract"
subtitle: "Heat Stress and Perinatal Outcomes in Ghana"
author: "Vincent Bio Bediako"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: false
      smooth_scroll: true
    number_sections: yes
    code_folding: show
    df_print: kable
    theme: flatly
    highlight: tango
    fig_caption: yes
  pdf_document:
    toc: yes
    toc_depth: 3
    number_sections: yes
    keep_tex: no
    latex_engine: xelatex
  github_document:
    toc: yes
    toc_depth: 3
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  message    = FALSE,
  warning    = FALSE,
  comment    = "#>",
  fig.align  = "center",
  fig.width  = 7,
  fig.height = 4.5,
  dpi        = 110
)
options(scipen = 999)
```

# Purpose and scope

This document implements steps (ii)–(iv) of the analytic plan accompanying the
proposal *Heat Stress and Perinatal Outcomes in Ghana: A Spatio-Temporal
Analysis* (Version 2.0). It produces the cleaned analytic spine that all
downstream modelling scripts will consume; it does not itself fit any
statistical model.

The three operations executed here are:

- **(ii) Deduplication.** The raw extract contains one row per maternal
  residency episode, which yields multiple rows per child when a mother's
  residency history spans more than one episode. The spine is reduced to one
  row per unique child, and the maternal residency episodes are retained in a
  separate long-format table that downstream survival or migration-history
  models can join back to.
- **(iii) Recoding of HDSS missing-data sentinels.** Per the NHRC value-label
  codebook (delivered 20 May 2026), the integers `8` and `9` denote
  "NaN/missing" and "DK" in the categorical fields. The codebook conventions
  are applied here. In addition, earlier correspondence flagged `88` and `99`
  as missing in `ceb`, `prgmonth`, and `numanc`; these are also recoded
  defensively, though neither appears in the present sample. The value `0`
  in `child_bwgt` denotes a child who was not weighed; biologically
  implausible weights (>6 kg or <0.5 kg) and the residual code `9.9` are
  also set to `NA`. Blank cells in `mother_mstatus` are recoded to `NA`.
- **(iv) Construction of derived analytic variables.** Mortality indicators
  (stillbirth-as-recorded, day-0, early-neonatal, neonatal, infant,
  under-five), birth-weight categories (LBW, VLBW), maternal-age and parity
  categories, season of delivery on both the three-season and two-season
  conventions for northern Ghana, reconstructed pregnancy windows for
  downstream DLNM linkage, and a flag for the **post-2012 ANC-reliable
  sub-cohort** used in the care-modifier sub-analyses.

**Variable name reconciliation.** The 20 May 2026 sample uses three field
names that differ from the 30 April 2026 sample: `cordres1` → `corddress`,
`delivery` → `delivery_type`, `gender` → `child_gender`. The script uses
the new names throughout; if it is re-run against the older extract these
will need to be renamed back.

**Analytic window.** Following the workflow agreed with NHRC, the primary
analytic window covers *all available birth-outcome records* in the full
extract, rather than a fixed ten-year span. This decision reflects the
relative scarcity of perinatal-mortality events and the value of preserving
the full longitudinal record for the main outcome models. The antenatal-care
variables (`numanc`, `prgmonth`, `ancatt`) are reliably populated only from
2013 onward; analyses that adjust for or stratify on ANC are therefore
restricted to a 2013-onward sub-cohort, with a binary flag (`anc_reliable`)
constructed in the spine for that purpose.

**Geocodes.** Household-level coordinates (`gps_latitude`, `gps_longitude`)
are present in this sample. Coverage is partial in the sample (concentrated
in more recent years); the full extract is expected to have broader
coverage. The spine preserves the coordinates as-is for downstream linkage
in `02_link_era5_utci.Rmd`.

The numeric category codes for `assist_del`, `child_birth_place`,
`delivery_type`, `cutcod`, `corddress`, `seventtype`, and `eeventtype`
will be converted to labelled factors in a follow-up labelling chunk
using the NHRC codebook delivered on 20 May 2026.

# Setup

```{r packages}
# Install any missing packages on first run:
# install.packages(c("haven","dplyr","tidyr","lubridate","stringr",
#                    "readr","knitr","kableExtra"))
library(haven)        # read_dta
library(dplyr)
library(tidyr)
library(lubridate)
library(stringr)
library(readr)
library(knitr)
library(kableExtra)

# Adjust these paths to match the project layout.
IN_PATH <- "vincent_phd_10data_20052026.dta"
OUT_DIR <- "."
dir.create(OUT_DIR, showWarnings = FALSE, recursive = TRUE)
```

A small helper used throughout the document to render diagnostic tables in a
consistent style:

```{r helper}
nice_table <- function(x, caption = NULL, full_width = FALSE,
                       wrap_col = NULL, wrap_width = "10cm") {
  # Detect output format (HTML vs LaTeX) so the styling adapts.
  fmt <- knitr::is_html_output()
  k <- kable(x, caption = caption, format.args = list(big.mark = ","),
             booktabs = TRUE, longtable = TRUE)
  if (fmt) {
    k <- kable_styling(k,
                       bootstrap_options = c("striped", "hover", "condensed"),
                       full_width = full_width, position = "left")
  } else {
    k <- kable_styling(k,
                       latex_options = c("striped", "hold_position", "repeat_header"),
                       full_width = full_width, position = "left")
  }
  # For tables with a wide description column, wrap that column at fixed width.
  if (!is.null(wrap_col)) {
    k <- column_spec(k, wrap_col, width = wrap_width)
  }
  k
}
```

# Load raw data

`haven::read_dta()` preserves the Stata variable labels and value labels as
attributes, so they remain accessible for documentation while we operate on
the underlying numeric codes.

```{r load-raw}
raw <- read_dta("C:/Users/bedia/Downloads/Chapter 3_Dissertation/vincent_phd_10data_20052026.dta")

tibble(metric = c("rows", "columns",
                  "unique mothers (motherid)",
                  "unique children (child_id)",
                  "unique compounds (locationid)"),
       value  = c(nrow(raw), ncol(raw),
                  dplyr::n_distinct(raw$motherid),
                  dplyr::n_distinct(raw$child_id),
                  dplyr::n_distinct(raw$locationid))) |>
  nice_table(caption = "Raw extract — basic dimensions")
```

The difference between raw rows and unique children reflects the
residency-episode duplication that step (ii) will resolve.

# Step (iii): Recode HDSS missing-data sentinels

Per the NHRC value-label codebook of 20 May 2026, the integers `8` (NaN) and
`9` (DK) denote missing values across the categorical fields. The values `88`
and `99` were flagged in earlier correspondence as missing in `ceb`,
`prgmonth`, and `numanc` and are recoded defensively, though neither appears
in the present sample. The value `0` in `child_bwgt` denotes a child who was
not weighed; biologically implausible weights and the residual code `9.9`
are also set to `NA`.

```{r recode}
sentinel_cat <- c(8, 9)
sentinel_num <- c(88, 99)

df <- raw |>
  mutate(
    ceb            = if_else(ceb %in% sentinel_num, NA_real_, ceb),
    prgmonth       = if_else(prgmonth %in% sentinel_num, NA_real_, prgmonth),
    numanc         = if_else(numanc %in% sentinel_num, NA_real_, numanc),
    across(c(assist_del, child_birth_place, delivery_type, cutcod,
             corddress, ancatt, first_nb, child_birth_type),
           ~ if_else(.x %in% sentinel_cat, NA_real_, .x)),
    child_bwgt     = if_else(child_bwgt == 0 |
                               child_bwgt == 9.9 |
                               child_bwgt < 0.5 |
                               child_bwgt > 6.0,
                             NA_real_, child_bwgt),
    mother_mstatus = as.character(mother_mstatus),
    mother_mstatus = if_else(str_trim(mother_mstatus) == "",
                             NA_character_, mother_mstatus)
  )
```

Effect of the recode on the non-missing counts of the affected fields:

```{r recode-summary}
before_vals <- c(
  ceb = sum(!is.na(raw$ceb)),
  prgmonth = sum(!is.na(raw$prgmonth)),
  numanc = sum(!is.na(raw$numanc)),
  child_bwgt = sum(!is.na(raw$child_bwgt)),
  mother_mstatus = sum(!is.na(as.character(raw$mother_mstatus)) &
                         trimws(as.character(raw$mother_mstatus)) != "")
)

after_vals <- c(
  ceb = sum(!is.na(df$ceb)),
  prgmonth = sum(!is.na(df$prgmonth)),
  numanc = sum(!is.na(df$numanc)),
  child_bwgt = sum(!is.na(df$child_bwgt)),
  mother_mstatus = sum(!is.na(df$mother_mstatus))
)

tibble(
  variable = names(before_vals),
  before_n = as.integer(before_vals),
  after_n = as.integer(after_vals),
  n_recoded_NA = before_n - after_n
) |>
  nice_table(caption = "Non-missing counts before and after sentinel recoding")
```

# Step (ii.a): Build the residency-episode table

Each maternal residency episode is kept on its own row in a separate
long-format table. This file is the join key for any analysis that needs
in-migration or out-migration timing.

```{r residency-episodes}
residency_episodes <- df |>
  distinct(motherid, sdate, edate, seventtype, eeventtype,
           locationid, location) |>
  arrange(motherid, sdate) |>
  mutate(episode_id = row_number(),
         sdate = as.Date(sdate),
         edate = as.Date(edate))

tibble(metric = c("episode rows", "unique mothers represented"),
       value  = c(nrow(residency_episodes),
                  dplyr::n_distinct(residency_episodes$motherid))) |>
  nice_table(caption = "Residency-episode table — dimensions")
```

A small preview of the table (sorted by mother and episode start date), to
confirm that the structure is as expected:

```{r residency-preview}
head(residency_episodes, 6) |>
  nice_table(caption = "First six rows of the residency-episode table")
```

# Step (ii.b): Deduplicate to one row per child

Where multiple rows share a `child_id`, the surviving row is chosen by three
ordered criteria. The first is **outcome preservation**: rows with a definite
birth outcome (`child_birth_type` = 1–5, covering abortion, miscarriage, live
birth, stillbirth, and neonatal death) take precedence over the
pregnancy-still-in-progress codes (6 = censored preg, 7 = still preg) and
over missing/DK codes. This avoids the failure mode in which stillbirths
and very-early neonatal deaths recorded at the time of the pregnancy event
are overwritten by later residency-only rows that carry a "censored
pregnancy" code. The second is lowest count of missing values across the
analytic covariate set. The third is tie-break on the most recent `sdate`.

```{r dedup}
analytic_vars <- c("mother_educ", "ses_index", "mother_mstatus",
                   "child_bwgt", "child_birth_place", "delivery_type",
                   "ancatt", "prgmonth", "numanc",
                   "first_nb", "assist_del", "cutcod", "corddress")

# Priority for outcome codes: definite outcomes (1-5) win over pregnancy
# in progress (6,7) which win over missing (NA).
outcome_priority <- function(x) {
  case_when(
    x %in% c(1, 2, 3, 4, 5) ~ 1L,   # abortion, miscarriage, live birth, SB, ND
    x == 6                  ~ 2L,   # censored preg
    x == 7                  ~ 3L,   # still preg
    TRUE                    ~ 9L    # missing/DK
  )
}

# ---------------------------------------------------------------------------
# Pregnancy-level identifier.
#
# child_id is populated only for LIVE births, so deduplicating on child_id
# silently collapses every non-live outcome (stillbirth, miscarriage,
# abortion), whose child_id is NA, into a single row — which is why a direct
# child_id-based spine under-counts stillbirths. For a perinatal / pregnancy-
# outcome analysis the unit of analysis must be the PREGNANCY, not the live
# child.
#
# Every row carries child_dob, the delivery date, which is present for non-
# live outcomes as well (the date the pregnancy ended in delivery). A
# pregnancy is therefore identified by the mother and the delivery date:
# distinct (motherid, child_dob) pairs are distinct pregnancies, capturing
# stillbirths and other non-live outcomes that child_id excludes. Multiple
# births (twins) share (motherid, child_dob) and collapse to one pregnancy,
# which is correct at the pregnancy level; the outcome-priority ordering
# below ensures the most severe / most complete outcome row represents such
# a pregnancy.
# ---------------------------------------------------------------------------
df <- df |>
  mutate(
    preg_id = if_else(
      !is.na(motherid) & !is.na(child_dob),
      paste0("P_", as.integer(factor(paste(motherid, as.integer(child_dob),
                                           sep = "_")))),
      NA_character_
    )
  )

spine_pre <- df |>
  mutate(.outcome_priority = outcome_priority(child_birth_type),
         .miss_score       = rowSums(is.na(across(all_of(analytic_vars))))) |>
  # Within each pregnancy, order so the definite/most-complete outcome row
  # is kept (e.g. a recorded stillbirth or live birth over a censored row),
  # mirroring the previous child-level priority but at the pregnancy level.
  arrange(preg_id, .outcome_priority, .miss_score, desc(sdate)) |>
  distinct(preg_id, .keep_all = TRUE) |>
  dplyr::select(-.outcome_priority, -.miss_score)

# Rows with no pregnancy key (motherid or child_dob missing) cannot be placed
# on the pregnancy timeline; quantify them rather than dropping silently.
n_no_pregkey <- sum(is.na(spine_pre$preg_id))

stopifnot(!anyDuplicated(spine_pre$preg_id[!is.na(spine_pre$preg_id)]))

tibble(metric = c("rows in spine (= unique pregnancies)",
                  "distinct mothers",
                  "live-birth pregnancies (child_birth_type == 3)",
                  "non-live outcomes (stillbirth/miscarriage/abortion/ND)",
                  "  of which recorded stillbirths (type 4)",
                  "pregnancies with a populated child_id",
                  "rows in raw extract",
                  "rows removed by deduplication",
                  "rows with no pregnancy key (motherid/child_dob NA)"),
       value  = c(nrow(spine_pre),
                  dplyr::n_distinct(spine_pre$motherid),
                  sum(spine_pre$child_birth_type == 3, na.rm = TRUE),
                  sum(spine_pre$child_birth_type %in% c(1, 2, 4, 5), na.rm = TRUE),
                  sum(spine_pre$child_birth_type == 4, na.rm = TRUE),
                  sum(!is.na(spine_pre$child_id)),
                  nrow(df),
                  nrow(df) - nrow(spine_pre),
                  n_no_pregkey)) |>
  nice_table(caption = "Effect of pregnancy-level deduplication (one row per mother × delivery date)")
```

**Why pregnancy-level, not child-level.** The previous spine deduplicated on
`child_id`. In this extract `child_id` is populated for non-live outcomes as
well as live births, so the loss of stillbirths under the old approach arose
not from missing identifiers but from the deduplication grouping: keying and
tie-breaking on `child_id` could discard the stillbirth row of a pregnancy in
favour of another row sharing that identifier. Keying instead on the
pregnancy (`motherid` × `child_dob`) groups rows by the actual pregnancy
event, so each recorded outcome—live birth, stillbirth, miscarriage, or
neonatal death—is preserved as its own pregnancy row. The non-live-outcome
counts above are the records the Stream A perinatal/early-neonatal analysis
depends on, and they are now retained in full (compare the birth-outcome
distribution table below).

**Interpreting the dedup count.** The rows removed by deduplication are not
"lost" data: they are redundant *maternal-context copies* of pregnancies that
appear on more than one residency row in the raw extract. Because the HDSS
data model is mother-centric (each row is a maternal residency episode,
not a pregnancy), a single pregnancy travels alongside the mother across each of
her recorded residency episodes, producing one row per (pregnancy, maternal
residency) pair. The dedup collapses these to one row per pregnancy for
modelling purposes, while the full residency-episode history is preserved
separately in `residency_episodes.rds` and can be joined back to the spine
on `motherid` for analyses that require maternal-migration context. No
pregnancy is dropped: the post-dedup row count equals the number of distinct
(`motherid`, `child_dob`) pregnancies in the raw extract, live and non-live
alike.

**A note on temporal snapshots in the spine.** Because the dedup tie-break
selects the most recent `sdate` among rows tied on outcome priority and
missingness, maternal characteristics that vary over time (`ses_index`,
`mother_educ`, `mother_mstatus`, residential `location`) in the spine
reflect the latest available maternal record rather than the value at the
time of the child's birth. For most fields this matters little — maternal
education and SES typically change slowly — but it is worth flagging in
the analytic write-up as a documented analytic choice, and may be a topic
for discussion with NHRC on whether a different convention is preferred.

# Step (iv): Derived analytic variables

Several conventions used in this block deserve explicit mention.

- Age at death is computed as integer days from `child_dob` to `child_dod`.
  Children without a recorded death date are coded as `NA` for age at death,
  and as `0` (not `NA`) for the binary mortality indicators, so that those
  indicators are NA-safe for downstream regression.
- The 280-day term-gestation assumption used to back-date pregnancy start is
  a working default. Sensitivity analyses would also fit results at 259 and
  294 days to bracket gestational-age error.
- Northern Ghana climatology follows the conventional three-season split:
  *dry-cool* (Nov–Mar), *hot-dry* (Apr–May), and *wet* (Jun–Oct). A
  two-level wet-vs-dry version is also retained for parsimonious models.
- Parity is taken from `ceb` where available. Because `ceb` is missing for
  a substantial fraction of records (approximately 45% in the 20 May 2026
  sample), a binary nulliparous proxy is also computed from the
  fully-populated `first_nb` field.

```{r derive}
spine <- spine_pre |>
  mutate(
    # ---- Dates ----
    mother_dob = as.Date(mother_dob),
    child_dob  = as.Date(child_dob),
    child_dod  = as.Date(child_dod),
    sdate      = as.Date(sdate),
    edate      = as.Date(edate),

    # ---- Age at death (days) ----
    age_death_days = as.integer(child_dod - child_dob),

    # ---- Mortality indicators (NA-safe) ----
    died_day0     = if_else(!is.na(age_death_days) & age_death_days == 0,         1L, 0L),
    died_enn      = if_else(!is.na(age_death_days) & age_death_days >= 0
                            & age_death_days <= 7,                                1L, 0L),
    died_neonatal = if_else(!is.na(age_death_days) & age_death_days >= 0
                            & age_death_days <= 28,                               1L, 0L),
    died_infant   = if_else(!is.na(age_death_days) & age_death_days >= 0
                            & age_death_days <= 365,                              1L, 0L),
    died_u5       = if_else(!is.na(age_death_days) & age_death_days >= 0
                            & age_death_days <= 1826,                             1L, 0L),

    # ---- Stream A primary outcome: stillbirth + ENN composite ----
    # Stillbirth as recorded in child_birth_type (code = 4)
    stillbirth = if_else(!is.na(child_birth_type) & child_birth_type == 4, 1L, 0L),
    # Perinatal mortality (Stream A primary): stillbirth OR death within
    # the first 7 days of life
    perinatal  = if_else(stillbirth == 1 | died_enn == 1, 1L, 0L),

    # ---- Birth-weight derivatives ----
    lbw  = if_else(!is.na(child_bwgt), as.integer(child_bwgt <  2.5), NA_integer_),
    vlbw = if_else(!is.na(child_bwgt), as.integer(child_bwgt <  1.5), NA_integer_),

    # ---- Temporal markers ----
    birth_year  = year(child_dob),
    birth_month = month(child_dob),
    season3 = case_when(
      birth_month %in% c(11, 12, 1, 2, 3) ~ "dry_cool",
      birth_month %in% c(4, 5)            ~ "hot_dry",
      birth_month %in% c(6, 7, 8, 9, 10)  ~ "wet",
      TRUE                                ~ NA_character_),
    season2 = if_else(birth_month %in% 6:10, "wet", "dry"),

    # ---- Maternal age and parity ----
    mat_age_yrs = as.numeric(child_dob - mother_dob) / 365.25,
    mat_age_cat = cut(mat_age_yrs,
                      breaks = c(-Inf, 19.999, 24, 29, 34, Inf),
                      labels = c("<20", "20-24", "25-29", "30-34", "35+"),
                      right  = TRUE),
    parity_cat = cut(ceb,
                     breaks = c(0, 1, 2, 4, Inf),
                     labels = c("nulliparous", "1", "2-4", "5+"),
                     include.lowest = TRUE, right = TRUE),
    nulliparous_proxy = if_else(first_nb == 1, 1L, 0L),

    # ---- Education collapsed ----
    educ_any_secondary = if_else(!is.na(mother_educ),
                                 as.integer(mother_educ >= 2),
                                 NA_integer_),

    # ---- Reconstructed pregnancy windows (280-day term gestation) ----
    preg_start_assumed = child_dob - days(280),
    preg_t3_start      = child_dob - days(98),
    preg_last30        = child_dob - days(30),
    preg_last7         = child_dob - days(7),

    # ---- ANC-reliable sub-cohort flag ----
    # The antenatal-care fields are populated reliably only for births from
    # 2013 onward; analyses that adjust for or stratify on ANC are restricted
    # to this sub-cohort. The PRIMARY analytic window uses all available
    # birth-outcome records (no upper or lower year restriction).
    anc_reliable = as.integer(birth_year >= 2013),

    # ---- Labelled birth outcome ----
    outcome_class = case_when(
      child_birth_type == 1 ~ "abortion",
      child_birth_type == 2 ~ "miscarriage",
      child_birth_type == 3 ~ "live_birth",
      child_birth_type == 4 ~ "stillbirth",
      child_birth_type == 5 ~ "neonatal_death_at_record",
      child_birth_type == 6 ~ "censored_pregnancy",
      child_birth_type == 7 ~ "ongoing_pregnancy",
      TRUE                  ~ NA_character_)
  )
```

# Diagnostics

## Mortality event counts

The full-file event counts, before any analytic-window restriction:

```{r mortality-full}
spine |>
  summarise(
    across(c(died_day0, died_enn, died_neonatal, died_infant, died_u5), sum)
  ) |>
  pivot_longer(everything(), names_to = "indicator", values_to = "events") |>
  mutate(definition = recode(indicator,
    died_day0     = "Day-0 death (death on day of birth)",
    died_enn      = "Early-neonatal death (0–7 days)",
    died_neonatal = "Neonatal death (0–28 days)",
    died_infant   = "Infant death (0–365 days)",
    died_u5       = "Under-five death (0–1,826 days)"
  )) |>
  dplyr::select(indicator, definition, events) |>
  nice_table(caption = "Mortality event counts, full sample (1995–2020)")
```

## Birth-outcome (child_birth_type) distribution

The mortality table above counts deaths derived from age at death. This block
tabulates the recorded birth-outcome code directly, so the stillbirth count
(`child_birth_type == 4`) and the missing share are both explicit. Sentinel
values 8 (NaN) and 9 (DK) were recoded to NA upstream, so the NA row captures
records with a missing or don't-know birth outcome.

```{r birth-type-distribution}
# Raw code distribution, including NA.
birth_type_tab <- spine |>
  count(child_birth_type, name = "n") |>
  mutate(
    label = recode(as.character(child_birth_type),
      "1" = "abortion", "2" = "miscarriage", "3" = "live_birth",
      "4" = "stillbirth", "5" = "neonatal_death_at_record",
      "6" = "censored_pregnancy", "7" = "ongoing_pregnancy",
      .missing = "NA (missing / DK)"),
    pct = round(100 * n / sum(n), 2)
  ) |>
  arrange(child_birth_type) |>
  dplyr::select(child_birth_type, label, n, pct)

nice_table(birth_type_tab,
           caption = "Distribution of recorded birth outcome (child_birth_type), full file, including missing.")

# Explicit stillbirth headline.
n_sb    <- sum(spine$stillbirth == 1, na.rm = TRUE)
n_total <- nrow(spine)
n_na    <- sum(is.na(spine$child_birth_type))
cat(sprintf("Recorded stillbirths (child_birth_type == 4): %s of %s records (%.3f%%)\n",
            format(n_sb, big.mark = ","), format(n_total, big.mark = ","),
            100 * n_sb / n_total))
cat(sprintf("child_birth_type missing (NA after sentinel recode): %s (%.2f%%)\n",
            format(n_na, big.mark = ","), 100 * n_na / n_total))
```

The labelled `outcome_class` is derived directly from `child_birth_type`, so
its distribution should agree with the table above; this cross-check confirms
the labelling step operated as intended.

```{r outcome-class-distribution}
spine |>
  count(outcome_class, name = "n") |>
  mutate(pct = round(100 * n / sum(n), 2)) |>
  arrange(desc(n)) |>
  nice_table(caption = "Distribution of labelled birth outcome (outcome_class), full file.")
```

## ANC-reliable sub-cohort (2013 onward)

The 2013-onward sub-cohort is the operative one for any analysis that requires
the ANC variables, since these fields are systematically blank for births
before 2012. The full analytic window — used for the primary perinatal
outcome models — places no upper or lower year restriction beyond what is
present in the full extract delivered by NHRC.

```{r sub-cohort}
sub <- spine |> filter(anc_reliable == 1)

tibble(
  metric = c("n children in ANC-reliable sub-cohort",
             "early-neonatal deaths (0–7 days)",
             "neonatal deaths (0–28 days)",
             "infant deaths (0–365 days)",
             "records with valid birth weight",
             "low birth weight (<2.5 kg)",
             "records with non-missing numanc",
             "records with valid parity (ceb)"),
  value  = c(nrow(sub),
             sum(sub$died_enn,       na.rm = TRUE),
             sum(sub$died_neonatal,  na.rm = TRUE),
             sum(sub$died_infant,    na.rm = TRUE),
             sum(!is.na(sub$child_bwgt)),
             sum(sub$lbw,            na.rm = TRUE),
             sum(!is.na(sub$numanc)),
             sum(!is.na(sub$ceb)))
) |>
  nice_table(caption = "Event and observation counts in the ANC-reliable (2013-onward) sub-cohort")
```

## Distribution checks

Season of delivery in the ANC-reliable (2013-onward) sub-cohort, on both the
three-season and the two-season conventions:

```{r season-dist}
sub |>
  count(season3, name = "n") |>
  arrange(desc(n)) |>
  nice_table(caption = "Season of delivery (three-season), ANC-reliable sub-cohort")

sub |>
  count(season2, name = "n") |>
  arrange(desc(n)) |>
  nice_table(caption = "Season of delivery (wet/dry), ANC-reliable sub-cohort")
```

Maternal age category, full file (a useful background distribution to keep in
mind when interpreting age-stratified outputs later):

```{r mat-age-dist}
spine |>
  count(mat_age_cat, name = "n") |>
  arrange(mat_age_cat) |>
  nice_table(caption = "Maternal age category, full file")
```

## Missing-data audit

The audit below reports, for every variable that enters the analytic plan,
the non-missing count in the spine, the overall percent missing, and the
percent missing within the pre-2013 and 2013-onward sub-cohorts. The audit
serves three purposes. First, it produces a single transparent record of
what is and is not measured. Second, the era-stratified columns expose
which variables are *structurally* missing (because the field was not
collected before a given year) versus those that are *operationally*
missing within an era (and therefore require analytic handling).
Third, the audit gives the analyst and the Centre a shared reference
point for any conversation about missing-data handling for the full
extract.

A note on apparent vs latent missingness. The ANC fields `prgmonth` and
`numanc` appear in the audit as essentially complete (≈100% non-missing),
but in the pre-2013 period the recorded values are typically `0` (a
placeholder for "not recorded") rather than `NA`. This is a *latent*
missingness pattern that the audit does not detect on its own; it is
addressed analytically by restricting all ANC-conditioned models to the
2013-onward sub-cohort using the `anc_reliable` flag, where the recorded
values reflect genuine ANC visit counts and pregnancy months at first
visit.

```{r missingness-audit}
audit_vars <- tribble(
  ~variable,            ~label,
  "gps_latitude",       "GPS latitude (household)",
  "gps_longitude",      "GPS longitude (household)",
  "mother_educ",        "Maternal education",
  "mother_mstatus",     "Maternal marital status",
  "ses_index",          "SES quintile",
  "child_bwgt",         "Birth weight (kg)",
  "ceb",                "Children ever born",
  "first_nb",           "First live birth indicator",
  "prgmonth",           "Pregnancy month at first ANC (see latent note)",
  "numanc",             "Number of ANC visits (see latent note)",
  "ancatt",             "ANC attendance",
  "child_birth_place",  "Place of delivery",
  "delivery_type",      "Mode of delivery",
  "assist_del",         "Birth attendance",
  "cutcod",             "Umbilical cord cutting",
  "corddress",          "Umbilical cord dressing"
)

audit <- audit_vars |>
  rowwise() |>
  mutate(
    n_nonmiss = sum(!is.na(spine[[variable]])),
    pct_miss  = round(mean(is.na(spine[[variable]])) * 100, 1),
    pct_miss_pre  = round(mean(is.na(spine[[variable]][spine$anc_reliable == 0])) * 100, 1),
    pct_miss_post = round(mean(is.na(spine[[variable]][spine$anc_reliable == 1])) * 100, 1)
  ) |>
  ungroup() |>
  dplyr::select(variable, label, n_nonmiss, pct_miss, pct_miss_pre, pct_miss_post)

audit |>
  nice_table(caption = "Missingness audit — non-missing counts and percent missing overall and by ANC-reliable era",
             wrap_col = 2, wrap_width = "7cm")
```

**Reading the audit.** The audit reveals four distinct missingness patterns
that drive the analytic handling decisions documented in Chapter 3
(§3.10.8). First, a set of variables (`mother_mstatus`, `mother_educ`,
maternal age, infant sex, place of delivery, delivery type, birth
attendance, cord-care fields) are essentially complete (<1% missing) and
require no special handling. Second, `child_bwgt` shows a strong era
gradient: about 81% missing before 2013 and about 30% missing thereafter,
reflecting the rise in facility deliveries and routine weighing over the
analytic window. Third, `gps_latitude` and `gps_longitude` are missing for
about 70% of records overall, with only a modest era gradient (72% vs
64%), indicating that the geocoding programme has not been retroactively
applied uniformly to the historical record. Fourth, `ceb` is missing for
about 39% of records across both eras, consistent with
respondent-conditioned missingness that affects parity reporting throughout
the surveillance period.

# Spine codebook

The variable list below documents the derived analytic spine. It is also
saved to `spine_codebook.csv` for reference.

```{r codebook}
spine_codebook <- tribble(
  ~variable,             ~description,
  "child_id",            "Child/record identifier (NHRC); populated for live and non-live outcomes in this extract. Not a reliable pregnancy key on its own — see preg_id.",
  "preg_id",             "Pregnancy identifier = mother x delivery date (motherid x child_dob). Unit of analysis for Stream A; one row per pregnancy across live and non-live outcomes.",
  "motherid",            "Unique mother identifier.",
  "locationid",          "Compound identifier.",
  "gps_latitude",        "Household-level GPS latitude (decimal degrees).",
  "gps_longitude",       "Household-level GPS longitude (decimal degrees).",
  "location",            "Rural/Urban indicator (1=rural, 2=urban).",
  "sdate",               "Start date of the maternal residency episode that the surviving row represents.",
  "edate",               "End date of the maternal residency episode that the surviving row represents.",
  "seventtype",          "Event type at residency start (1=birth, 2=enter, 3=enumeration; NHRC codebook).",
  "eeventtype",          "Event type at residency end (1=death, 2=exited, 3=present; NHRC codebook).",
  "child_dob",           "Date of birth (delivery).",
  "child_dod",           "Date of death; NA if alive at last record.",
  "age_death_days",      "Integer days from child_dob to child_dod; NA if alive.",
  "died_day0",           "1 if death on day of birth.",
  "died_enn",            "1 if death within 0–7 days of birth (early neonatal).",
  "died_neonatal",       "1 if death within 0–28 days (neonatal).",
  "died_infant",         "1 if death within 0–365 days (infant).",
  "died_u5",             "1 if death within 0–1,826 days (under-five).",
  "stillbirth",          "1 if child_birth_type = 4 (stillbirth as recorded).",
  "perinatal",           "Stream A primary outcome: 1 if stillbirth OR died_enn.",
  "child_gender",        "Sex of child (1=male, 2=female; codebook applies).",
  "child_bwgt",          "Birth weight in kg; 0, 9.9, and outliers set to NA.",
  "lbw",                 "1 if birth weight < 2.5 kg; NA if weight missing.",
  "vlbw",                "1 if birth weight < 1.5 kg; NA if weight missing.",
  "child_birth_place",   "Place of birth (1=hospital, 2=health centre, 3=home, 4=tba home, 5=other; NHRC codebook).",
  "delivery_type",       "Mode of delivery (1=normal vaginal, 2=cesarean, 3=vacuum, 4=forceps, 5=vaginal cutting, 7=other; NHRC codebook).",
  "assist_del",          "Delivery assistance (1=doctor, 2=midwife, 3=nurse, 4=CHO, 5=TBA, 6=family, 7=other; NHRC codebook).",
  "cutcod",              "Material used to cut umbilical cord (1=new razor, 2=old razor, 3=knife, 4=sharp stone, 5=scissors, 6=other; NHRC codebook).",
  "corddress",           "Material used to dress umbilical stump (1=shea butter, 2=black powder + shea, 3=western medicine, 4=nothing, 5=other; NHRC codebook).",
  "ancatt",              "ANC attendance (1=yes, 2=no); reliable only 2013+.",
  "prgmonth",            "Pregnancy month at first ANC visit; reliable only 2013+.",
  "numanc",              "Number of ANC visits during pregnancy; reliable only 2013+.",
  "first_nb",            "1 if first live birth of mother; 2 otherwise.",
  "ceb",                 "Children ever born.",
  "parity_cat",          "Parity categories from ceb; NA where ceb missing.",
  "nulliparous_proxy",   "1 if first_nb==1 (use when ceb missing).",
  "mother_dob",          "Mother date of birth.",
  "mat_age_yrs",         "Maternal age at delivery (years).",
  "mat_age_cat",         "Maternal-age categories: <20, 20-24, 25-29, 30-34, 35+.",
  "mother_educ",         "Maternal education (0=none, 1=primary, 2=JHS, 3=SHS, 4=tertiary; NHRC codebook).",
  "educ_any_secondary",  "1 if mother_educ >= 2 (JHS or higher).",
  "mother_mstatus",      "Marital status (1=divorced, 2=died, 3=married, 4=separated, 5=single, 6=widowed; NHRC codebook).",
  "ses_index",           "SES quintile (1=poorest – 5=least poor; NHRC codebook).",
  "birth_year",          "Calendar year of birth.",
  "birth_month",         "Calendar month of birth.",
  "season3",             "Season of delivery (dry_cool / hot_dry / wet).",
  "season2",             "Season of delivery (dry / wet).",
  "preg_start_assumed",  "Assumed pregnancy start (child_dob - 280 days).",
  "preg_t3_start",       "Start of third trimester (child_dob - 98 days).",
  "preg_last30",         "30 days before delivery.",
  "preg_last7",          "7 days before delivery.",
  "anc_reliable",        "1 if birth in 2013 onward (ANC-reliable sub-cohort).",
  "child_birth_type",    "Raw integer outcome code (1=abortion, 2=miscarriage, 3=live birth, 4=stillbirth, 5=neonatal death, 6=censored preg, 7=still preg; 8/9 recoded NA per NHRC codebook).",
  "outcome_class",       "Labelled child_birth_type (abortion / miscarriage / live_birth / stillbirth / neonatal_death_at_record / censored_pregnancy / ongoing_pregnancy)."
)

nice_table(spine_codebook, caption = "Spine codebook",
           wrap_col = 2, wrap_width = "11cm")
```

# Persist outputs

The spine and the residency-episode table are written in both R-native
(`.rds`) and text (`.csv`) formats so that downstream scripts in R, Stata,
Python, or any other environment can read them without modification.

```{r save, results='hold'}
saveRDS(spine,              file.path(OUT_DIR, "spine.rds"))
saveRDS(residency_episodes, file.path(OUT_DIR, "residency_episodes.rds"))
write_csv(spine,              file.path(OUT_DIR, "spine.csv"))
write_csv(residency_episodes, file.path(OUT_DIR, "residency_episodes.csv"))
write_csv(spine_codebook,     file.path(OUT_DIR, "spine_codebook.csv"))

cat("Files written to ", normalizePath(OUT_DIR), ":\n  ",
    paste(c("spine.rds", "residency_episodes.rds",
            "spine.csv", "residency_episodes.csv",
            "spine_codebook.csv"),
          collapse = "\n  "), sep = "")
```

# Expected diagnostic counts (sample dataset, 20 May 2026)

The script has been validated against the second sample extract delivered
on 20 May 2026 (22,340 rows out of approximately 223,440 in the full
extract, covering birth outcomes and maternal residency from 1995 to 2020).
A successful run on that sample should reproduce the following counts; any
discrepancy indicates that something has changed in the input file and
should be investigated before downstream modelling proceeds. The full-data
run, performed by the Centre's analysts under the agreed workflow, will
produce substantially larger figures; the expected-counts table is
preserved as a sample-stage validation reference, not as a target for the
full data.

```{r expected, echo=FALSE}
tibble(
  metric = c("rows in raw extract",
             "unique children (spine)",
             "unique mothers",
             "unique compounds (locationid)",
             "distinct household GPS coordinates",
             "records with non-missing GPS",
             "birth-year range (sample)",
             "recorded stillbirths (child_birth_type = 4)",
             "early-neonatal deaths (0–7 days, full file)",
             "neonatal deaths (0–28 days, full file)",
             "perinatal mortality composite (SB + ENN)",
             "n children in ANC-reliable sub-cohort (2013+)",
             "valid birth weights (0.5–6.0 kg)",
             "LBW (<2.5 kg)"),
  expected_value = c("22,340", "8,876", "5,025", "2,885",
                     "904", "2,696 (~30%)",
                     "1995–2020",
                     "1", "122", "173", "123",
                     "2,942", "3,204", "347")
) |>
  nice_table(caption = "Expected diagnostic counts for the 20 MAY 2026 SAMPLE. Full-data figures will differ.",
             wrap_col = 1, wrap_width = "10cm")
```

# Session information

```{r session}
sessionInfo()
```

# Next steps

With the spine in hand, the immediate items are, in order:

1. **Apply the NHRC value-label codebook** (received 20 May 2026) for
   `assist_del`, `child_birth_place`, `delivery_type`, `cutcod`,
   `corddress`, `mother_mstatus`, `child_gender`, `location`, `ses_index`,
   `mother_educ`, `child_birth_type`, `seventtype`, and `eeventtype`, and
   add a labelling chunk to this document that promotes these integer
   fields to labelled factors.
2. Write `02_link_era5_utci.Rmd`, which (i) reads `spine.rds`,
   (ii) consumes the household-level GPS coordinates that are already in
   the spine, (iii) links each pregnancy to a daily UTCI series from
   ERA5-Land at the household coordinate, and (iv) writes the
   pregnancy-level exposure matrix to `exposure.rds`. Per the workflow
   agreed with NHRC, all climate-variable generation — ERA5-Land download,
   UTCI computation following Di Napoli et al. (2018), and
   gestational-window aggregation — is conducted within this script and
   executed by the Centre's analysts against the full extract.
3. Write `03_phase1_descriptive.Rmd`, `04_phase2_misclassification.Rmd`,
   `05_phase3_dlnm_perinatal.Rmd`, `06_phase4_casecrossover.Rmd`,
   `07_phase5_dlnm_birthweight.Rmd`, `08_phase6_effect_modification.Rmd`,
   and `09_phase7_sensitivity.Rmd`, in line with the Chapter 3 analytic
   plan. The effect-modification and ANC-related analyses use the
   2013-onward `anc_reliable` sub-cohort; the primary perinatal-outcome
   models use the full birth-year range available in the NHRC extract.
