---
title: "Linking Daily UTCI Exposure to Pregnancies — Navrongo HDSS"
subtitle: "Heat Stress and Perinatal Outcomes in Ghana"
author: "Vincent"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
    df_print: paged
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  comment = "#>"
)

# Defaults so conditional chunk eval= never fails before the config chunk runs.
ERA5_ACQUISITION     <- "cds_download"
UTCI_SOURCE          <- "local_ecmwf_daily"
RUN_LOCAL_DAILY      <- TRUE
RUN_CDS_DOWNLOAD     <- FALSE
RUN_UTCI_PRODUCT     <- FALSE
RUN_UTCI_COMPUTE     <- FALSE
RUN_UTCI_PRECOMPUTED <- TRUE
```

# Purpose and scope

This document implements the climate-linkage step of the analytic plan
accompanying Chapter 3 (Methodology, Version 8.0). It consumes the
pregnancy-level spine produced by `01_build_spine.Rmd` and produces a
pregnancy-level **daily UTCI exposure matrix** for the 0–30 day window
pre-delivery, together with trimester and outcome-anchored summaries that
feed downstream DLNM, case-crossover, and sensitivity analyses.

Three operations are executed here.

- **(i) ERA5-Land input acquisition.** Four hourly variables (2 m air
  temperature, 2 m dewpoint temperature, 10 m wind speed, downward
  shortwave radiation) are obtained for the Navrongo catchment bounding box
  across the 1995–2020 analytic window. The acquisition method is
  configurable: download-at-runtime via the Copernicus CDS API, or
  read-from-disk if the inputs are already staged.
- **(ii) Daily UTCI exposure.** The exposure is the **daily maximum UTCI**
  from ECMWF's pre-computed daily product (the same product used in
  Chapter 2, read here at daily resolution). Each file is one NetCDF per day
  carrying `utci_daily_min` and `utci_daily_max`; only the maximum is used.
  Mean radiant temperature is computed by ECMWF from the full ERA5 radiation
  fields, so the exposure does not depend on any in-script radiation
  modelling. The source is configurable, but the default reads the daily
  files already staged on disk.
- **(iii) Pregnancy-level exposure assembly.** For each pregnancy in the
  spine, the script extracts the daily UTCI series for the 0–30 day window
  pre-delivery (the DLNM lag matrix), the trimester means (T1, T2, T3,
  full-pregnancy), and the outcome-anchored short-window means (last 7,
  14, 30 days). The resulting pregnancy-level exposure matrix is written
  to `exposure.rds` for downstream DLNM and case-crossover scripts.

**Note on workflow.** Per the code-then-execute arrangement agreed with
NHRC, the researcher develops and tests this script against a sample
dataset and stages the inputs for NHRC analysts to execute against the
full extract on the Centre's secure infrastructure. The configuration
block below exposes all environment-dependent paths and credentials so
that the script runs identically on the development environment and on
the Centre's compute environment.

# Configuration

The configuration block below exposes the operational choices flagged in
the chapter discussion of `02_link_era5_utci.Rmd`. Three of these reflect
unresolved decisions that may be settled in conversation with NHRC; the
script is written to support each option without restructuring.

```{r config}
# ============================================================================
# Paths
# ============================================================================
IN_PATH      <- "spine.rds"                # output of 01_build_spine.Rmd
OUT_DIR      <- "."                         # where exposure.rds is written

# Directory holding the downloaded ECMWF daily UTCI files, one NetCDF per day,
# named ECMWF_utci_daily_stats_YYYYMMDD_v1.1_con*.nc, each with the layers
# utci_daily_min and utci_daily_max. This is the same product used in
# Chapter 2, read here at daily resolution for the DLNM lag matrix.
UTCI_DIR     <- "C:/Users/bedia/Downloads/Chapter 3_Dissertation"

CLIMATE_DIR  <- "era5_land_navrongo"        # cache for the in-script ERA5 path (legacy)
LOG_DIR      <- file.path(OUT_DIR, "logs")  # extraction diagnostics

dir.create(CLIMATE_DIR, showWarnings = FALSE, recursive = TRUE)
dir.create(LOG_DIR,     showWarnings = FALSE, recursive = TRUE)

# ============================================================================
# Catchment bounding box (Kassena-Nankana districts, Upper East Region)
# Slight padding on each side to ensure no household falls on a tile edge.
# ============================================================================
BBOX <- list(
  lat_min = 10.50, lat_max = 11.20,
  lon_min = -1.45, lon_max = -0.75
)

# ============================================================================
# Analytic window (matches Chapter 3 §3.8.1)
# ============================================================================
YEAR_MIN <- 1995L
YEAR_MAX <- 2020L

# ============================================================================
# Exposure construction parameters (matches Chapter 3 §3.10.3)
# ============================================================================
LAG_MIN          <- 0L    # day of delivery
LAG_MAX          <- 30L   # 30 days before delivery
GESTATION_DAYS   <- 280L  # term gestation; sensitivity at 259 and 294
TRIMESTER1_END   <- 91L   # days 0–91 post-conception
TRIMESTER2_END   <- 182L  # days 92–182
# Trimester 3 = days 183–280 (delivery)
OUTCOME_WINDOWS  <- c(7L, 14L, 30L)  # last-N-day means

# ============================================================================
# Daily UTCI summary statistic
# Chapter 3 is an acute-heat-stress study: the exposure is the daily MAXIMUM
# UTCI throughout. The ECMWF daily files carry utci_daily_min and
# utci_daily_max; only the maximum is used. The percentile and heatwave
# metrics below are defined on the same daily-maximum distribution, so the
# whole exposure apparatus is internally consistent.
# ============================================================================
DAILY_STAT_PRIMARY     <- "max"   # daily maximum UTCI — the sole exposure

# ============================================================================
# Percentile / heatwave exposure metrics (carried over from Chapter 2).
# Location-specific percentiles: each grid cell's own UTCI distribution over
# the analytic window defines its p90/p95 thresholds. A heatwave run is a
# consecutive spell of days at or above the p90 threshold.
# ============================================================================
PERCENTILES        <- c(90, 95)   # location-specific exposure percentiles
HEATWAVE_PCTL      <- 90          # threshold percentile for run-length metric
HEATWAVE_MIN_RUN   <- 2L          # minimum consecutive days to count as a run

# ============================================================================
# OPERATIONAL CHOICES — to be confirmed with NHRC climate-expertise lead
# ============================================================================
#
# Choice 1: UTCI computation source. A first run computing MRT in-script from
# ERA5-Land's downward-shortwave flux alone produced daily-MAXIMUM UTCI that
# was implausibly high (median ~35 °C across all seasons, with almost no
# seasonal contrast). The cause is that MRT cannot be computed correctly from
# ERA5-Land alone: a defensible mean-radiant-temperature calculation (Di
# Napoli et al. 2018) needs the direct-beam shortwave (fdir) and the net
# short- and long-wave fluxes (ssr, str, strd), which ERA5-Land does not
# provide in the form the calculation requires. ECMWF already computed UTCI
# correctly from full ERA5 and published it as a ready-to-use daily product;
# reading that product is therefore both more correct and less code than any
# in-script reconstruction from ERA5-Land. The options below are ordered from
# most to least recommended:
#
# Choice 1: UTCI source. The default reads the ECMWF pre-computed daily UTCI
# files already downloaded for Chapter 2 (one NetCDF per day, with the layers
# utci_daily_min and utci_daily_max). These embed the correct mean radiant
# temperature computed by ECMWF from the full ERA5 radiation fields, so the
# inflation seen in an earlier in-script run (median ~35 °C across all
# seasons, no seasonal contrast) does not arise. The options below are
# ordered from most to least recommended:
#
#   "local_ecmwf_daily"         — read the local ECMWF daily UTCI files in
#                                 UTCI_DIR (RECOMMENDED; the default). One
#                                 NetCDF per day; correct MRT already embedded.
#                                 This is the same product Chapter 2 used.
#   "cds_utci_product"          — download ECMWF daily UTCI from the CDS
#                                 "derived-utci-historical" dataset if the
#                                 local files are not present.
#   "compute_in_script_full"    — compute MRT via the full Di Napoli (2018)
#                                 radiation balance; REQUIRES the full ERA5
#                                 single-levels radiation variables, NOT
#                                 ERA5-Land. Use only if those inputs are
#                                 staged. See the MRT section for caveats.
#   "compute_in_script_package" — DEPRECATED. ERA5-Land + simplified MRT.
#                                 Retained only for reproducing the first run;
#                                 known to inflate daily-max UTCI. Do not use
#                                 for analysis.
UTCI_SOURCE <- "local_ecmwf_daily"

ERA5_ACQUISITION <- "cds_download"

# Chunk flags consumed by knitr eval= options (updated here after paths/modes set).
RUN_LOCAL_DAILY      <- (UTCI_SOURCE == "local_ecmwf_daily")
RUN_CDS_DOWNLOAD     <- (ERA5_ACQUISITION == "cds_download" &&
                          UTCI_SOURCE %in% c("compute_in_script_full",
                                             "compute_in_script_package"))
RUN_UTCI_PRODUCT     <- (UTCI_SOURCE == "cds_utci_product")
RUN_UTCI_COMPUTE     <- (UTCI_SOURCE %in% c("compute_in_script_full",
                                            "compute_in_script_package"))
RUN_UTCI_PRECOMPUTED <- (UTCI_SOURCE %in% c("local_ecmwf_daily",
                                            "cds_utci_product"))

# ============================================================================
# Copernicus CDS credentials (used only if ERA5_ACQUISITION == "cds_download")
#
# Configure once per machine — never paste a token into this file or Git:
#   • Set env var CDS_API_KEY (recommended), e.g. in .Renviron:
#       CDS_API_KEY=your-personal-access-token
#   • Or run interactively once per session:
#       ecmwfr::wf_set_key(key = "your-personal-access-token")
#
# Obtain a PAT from https://cds.climate.copernicus.eu/profile after
# accepting the ERA5-Land licence terms.
# ============================================================================
```

```{r libraries}
# Core packages required regardless of configuration. Install on first
# environment setup with:
#   install.packages(c("dplyr","tidyr","lubridate","purrr","readr",
#                      "stringr","terra","knitr","kableExtra"))
#
# Configuration-conditional packages 
#   ERA5_ACQUISITION = "cds_download"  → install.packages("ecmwfr")
#   UTCI_SOURCE      = "compute_in_script_package" → install.packages("comf")

suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(lubridate)
  library(purrr)
  library(readr)
  library(stringr)
  library(terra)
  library(knitr)
  library(kableExtra)
})

# Conditionally-loaded packages: only required for specific configurations.
# Loading them on demand keeps the script runnable in environments that
# only need a subset of capabilities.
load_if_needed <- function(pkg) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    stop("Package '", pkg, "' is required for the chosen configuration but is not installed. ",
         "Install with: install.packages('", pkg, "')")
  }
  suppressPackageStartupMessages(library(pkg, character.only = TRUE))
}

if (ERA5_ACQUISITION == "cds_download")           load_if_needed("ecmwfr")
if (UTCI_SOURCE      == "compute_in_script_package") load_if_needed("comf")
```

```{r helpers}
nice_table <- function(x, caption = NULL, full_width = FALSE,
                       wrap_col = NULL, wrap_width = "10cm") {
  fmt <- knitr::is_html_output()
  k <- kable(x, caption = caption, format.args = list(big.mark = ","),
             booktabs = TRUE, longtable = TRUE)
  if (fmt) {
    k <- kable_styling(k,
                       bootstrap_options = c("striped", "hover", "condensed"),
                       full_width = full_width, position = "left")
  } else {
    k <- kable_styling(k,
                       latex_options = c("striped", "hold_position", "repeat_header"),
                       full_width = full_width, position = "left")
  }
  if (!is.null(wrap_col)) k <- column_spec(k, wrap_col, width = wrap_width)
  k
}

# Defensive coordinate check: catches NA, infinite, and any value outside
# the catchment bounding box (a sentinel like 0/0 or a swapped lat/lon).
valid_coords <- function(lon, lat,
                         lat_min = BBOX$lat_min, lat_max = BBOX$lat_max,
                         lon_min = BBOX$lon_min, lon_max = BBOX$lon_max) {
  is.finite(lon) & is.finite(lat) &
    lon >= lon_min & lon <= lon_max &
    lat >= lat_min & lat <= lat_max
}

# --- Shared climate helpers (needed by both the UTCI-product path and the
# in-script ERA5-Land path, so defined here in an always-run chunk) ---------

# Configure Copernicus CDS access from an environment variable. Never store
# the token in the script. Set CDS_API_KEY in the session or the .Renviron.
configure_cds <- function() {
  load_if_needed("ecmwfr")
  key <- Sys.getenv("CDS_API_KEY", unset = "")
  if (nzchar(key)) ecmwfr::wf_set_key(key = key)
}

# Robustly parse the hourly time axis from a CDS NetCDF SpatRaster. CDS files
# often encode valid_time in the layer names rather than a GDAL-readable time
# dimension, so terra::time() can return NA.
parse_era5_times <- function(spat_rast) {
  t <- terra::time(spat_rast)
  if (length(t) == terra::nlyr(spat_rast) && !all(is.na(t))) return(t)
  nm <- names(spat_rast)
  if (any(grepl("valid_time=", nm, fixed = TRUE))) {
    unix_sec <- as.numeric(sub(".*valid_time=", "", nm))
    return(as.POSIXct(unix_sec, origin = "1970-01-01", tz = "UTC"))
  }
  stop(sprintf("Could not determine time axis for %s",
               basename(terra::sources(spat_rast)[1])))
}

# Cosine of the solar zenith angle at given coordinates and UTC times,
# via standard astronomical formulae (Michalsky 1988 simplified). Returns a
# locations × times matrix. Used by the full Di Napoli (2018) MRT path.
solar_cos_zenith <- function(lat, lon, times_utc) {
  lat_r <- lat * pi / 180
  # Day-of-year and fractional UTC hour
  doy  <- as.numeric(format(times_utc, "%j"))
  hutc <- as.numeric(format(times_utc, "%H")) +
          as.numeric(format(times_utc, "%M")) / 60
  # Solar declination (radians)
  decl <- 0.409 * sin(2 * pi / 365 * doy - 1.39)
  # Equation of time (minutes) → hour-angle correction
  B   <- 2 * pi * (doy - 81) / 364
  eot <- 9.87 * sin(2 * B) - 7.53 * cos(B) - 1.5 * sin(B)
  # Build the locations × times matrix
  n_loc <- length(lat); n_t <- length(times_utc)
  out <- matrix(NA_real_, n_loc, n_t)
  for (j in seq_len(n_t)) {
    # Local solar time hour angle (degrees → radians); lon east-positive
    tst <- hutc[j] * 60 + eot[j] + 4 * lon
    ha  <- ((tst / 4) - 180) * pi / 180
    out[, j] <- sin(lat_r) * sin(decl[j]) +
                cos(lat_r) * cos(decl[j]) * cos(ha)
  }
  out
}

# --- Helpers carried over from the Chapter 2 daily pipeline, adapted to read
# the ECMWF daily UTCI files (one NetCDF per day, layers utci_daily_min and
# utci_daily_max). These are proven against the actual downloaded files. ------

# Parse the calendar date from an ECMWF daily filename. The date key is taken
# from the filename (e.g. ..._stats_19950115_v1.1_...) rather than the
# embedded time dimension, which is more robust to malformed time axes.
parse_date_from_filename <- function(filepath) {
  x  <- basename(filepath)
  m1 <- regmatches(x, regexpr("(19|20)\\d{2}(0[1-9]|1[0-2])(0[1-9]|[12]\\d|3[01])", x))
  if (length(m1) == 1 && nchar(m1) == 8) return(as.Date(m1, "%Y%m%d"))
  m2 <- regmatches(x, regexpr("(19|20)\\d{2}[-_](0[1-9]|1[0-2])[-_](0[1-9]|[12]\\d|3[01])", x))
  if (length(m2) == 1 && nchar(m2) == 10) {
    return(as.Date(paste0(substr(m2,1,4), "-", substr(m2,6,7), "-", substr(m2,9,10))))
  }
  as.Date(NA)
}

# Select the requested daily statistic from an ECMWF daily SpatRaster.
# The files carry utci_daily_min and utci_daily_max. Chapter 3 uses "max"
# (the sole exposure). The "min" and midrange options are provided only for
# completeness; the midrange (min+max)/2 is NOT the true daily mean and is
# not used as an exposure anywhere in this pipeline.
get_daily_layer <- function(r, metric = DAILY_STAT_PRIMARY) {
  nms <- names(r)
  if (all(c("utci_daily_min", "utci_daily_max") %in% nms)) {
    rmin <- r[["utci_daily_min"]]
    rmax <- r[["utci_daily_max"]]
    if (metric == "min")      return(rmin)
    if (metric == "max")      return(rmax)
    if (metric == "midrange") return((rmin + rmax) / 2)  # not a daily mean
  }
  if (any(grepl("utci", nms, ignore.case = TRUE))) {
    return(r[[which(grepl("utci", nms, ignore.case = TRUE))[1]]])
  }
  r[[1]]
}

# Some CDS downloads arrive as zip-disguised NetCDFs. Detect and repair them
# so terra::rast() can open them.
is_zipfile <- function(f) {
  con <- file(f, "rb"); on.exit(close(con))
  sig <- readBin(con, what = "raw", n = 2)
  identical(sig, as.raw(c(0x50, 0x4B)))
}

fix_zip_nc <- function(f, out_dir = tempdir()) {
  td <- tempfile("unz_"); dir.create(td)
  on.exit(unlink(td, recursive = TRUE), add = TRUE)
  utils::unzip(f, exdir = td)
  inner <- list.files(td, pattern = "\\.nc$", full.names = TRUE, recursive = TRUE)
  if (length(inner) == 0) { warning("Zip had no .nc inside: ", f); return(NA_character_) }
  dest <- file.path(out_dir, basename(f))
  file.copy(inner[1], dest, overwrite = TRUE)
  dest
}

# Detect whether a layer is in Kelvin (ECMWF UTCI ships in K) and needs
# conversion to °C.
is_kelvin <- function(r_layer) {
  v <- terra::values(r_layer); v <- v[!is.na(v)]
  if (length(v) == 0) return(FALSE)
  mean(v[seq_len(min(500, length(v)))], na.rm = TRUE) > 100
}
```

# Load the spine

```{r load-spine}
#Directory need to be changed
spine <- readRDS("C:/Users/bedia/Downloads/Chapter 3_Dissertation/spine.rds")

cat(sprintf("Spine rows (= unique children): %s\n", format(nrow(spine), big.mark = ",")))
cat(sprintf("Birth-year range: %d–%d\n",
            min(spine$birth_year, na.rm = TRUE),
            max(spine$birth_year, na.rm = TRUE)))
cat(sprintf("Distinct mothers: %s\n", format(dplyr::n_distinct(spine$motherid), big.mark = ",")))
cat(sprintf("Distinct compounds: %s\n", format(dplyr::n_distinct(spine$locationid), big.mark = ",")))
```

## Coverage gate — geocodes

Pregnancies with valid household coordinates enter the household-coordinate
exposure linkage; the residual subset is held for the community-centroid
sensitivity branch (§3.10.7, item i in Chapter 3).

```{r gps-coverage}
spine <- spine |>
  mutate(coord_valid = valid_coords(gps_longitude, gps_latitude))

gps_summary <- tibble(
  metric = c("rows in spine",
             "valid household coordinates",
             "coordinates outside catchment bbox",
             "missing coordinates",
             "valid coordinate coverage (%)"),
  value  = c(nrow(spine),
             sum(spine$coord_valid, na.rm = TRUE),
             sum(!is.na(spine$gps_latitude) & !spine$coord_valid),
             sum(is.na(spine$gps_latitude) | is.na(spine$gps_longitude)),
             round(100 * mean(spine$coord_valid, na.rm = TRUE), 2))
)

nice_table(gps_summary, caption = "GPS coverage diagnostics")

# Primary exposure linkage uses only the geocoded subset.
preg <- spine |>
  filter(coord_valid,
         birth_year >= YEAR_MIN, birth_year <= YEAR_MAX) |>
  dplyr::select(child_id, motherid, locationid,
                gps_longitude, gps_latitude,
                child_dob, preg_start_assumed,
                preg_t3_start, preg_last30, preg_last7) |>
  rename(lon = gps_longitude, lat = gps_latitude) |>
  # Defensive coercion: spine.rds may store dates as Date, POSIXct, or
  # character depending on the upstream version. Coerce explicitly and
  # drop any row where either the delivery date or the reconstructed
  # pregnancy start is missing (these break seq.Date() downstream).
  mutate(
    child_dob          = as.Date(child_dob),
    preg_start_assumed = as.Date(preg_start_assumed),
    preg_t3_start      = as.Date(preg_t3_start),
    preg_last30        = as.Date(preg_last30),
    preg_last7         = as.Date(preg_last7)
  ) |>
  filter(!is.na(child_dob), !is.na(preg_start_assumed)) |>
  mutate(row_id = row_number())

cat(sprintf("\nPregnancies entering household-coordinate exposure linkage: %s\n",
            format(nrow(preg), big.mark = ",")))
```

## Unique household locations

Many pregnancies share a household coordinate. Extracting at the
location-level (rather than the pregnancy level) makes the climate-linkage
step approximately *k*-fold faster, where *k* is the average pregnancies
per location. The exposure values are joined back to the pregnancy table
via the `loc_id` index.

```{r unique-locations}
locations <- preg |>
  distinct(lon, lat) |>
  arrange(lat, lon) |>
  mutate(loc_id = row_number())

preg <- preg |>
  left_join(locations, by = c("lon", "lat"))

cat(sprintf("Unique household coordinates: %s\n",
            format(nrow(locations), big.mark = ",")))
cat(sprintf("Mean pregnancies per location: %.1f\n",
            nrow(preg) / nrow(locations)))
```

# Build the (location, day) extraction grid

For each pregnancy, the analytic window of interest is the union of:
the daily DLNM lag window (`child_dob - 30 days` to `child_dob`); the
trimester windows (`preg_start_assumed` to `child_dob`); and the
outcome-anchored short windows (already covered by the DLNM window).
The trimester windows extend further back in time than the DLNM
window. The script builds the **union** of these as the (location, day)
extraction grid.

```{r needed-days}
preg_windows <- preg |>
  mutate(
    win_start = pmin(preg_start_assumed, child_dob - days(LAG_MAX)),
    win_end   = child_dob
  ) |>
  dplyr::select(row_id, loc_id, win_start, win_end)

# Expand to one row per (pregnancy, day)
preg_days <- preg_windows |>
  rowwise() |>
  mutate(day = list(seq.Date(win_start, win_end, by = "day"))) |>
  ungroup() |>
  unnest(day) |>
  dplyr::select(row_id, loc_id, day)

# (location, day) extraction grid — what climate data must be read
extract_grid <- preg_days |>
  distinct(loc_id, day) |>
  arrange(loc_id, day)

cat(sprintf("Total (pregnancy, day) pairs needed: %s\n",
            format(nrow(preg_days), big.mark = ",")))
cat(sprintf("Unique (location, day) pairs to extract: %s\n",
            format(nrow(extract_grid), big.mark = ",")))
cat(sprintf("Compression ratio: %.2fx\n",
            nrow(preg_days) / nrow(extract_grid)))

# Save the extraction grid to disk; the UTCI step reads this to limit
# month-tiles, locations, and calendar days processed.
saveRDS(extract_grid, file.path(LOG_DIR, "extract_grid.rds"))
```

# ERA5-Land input acquisition

This section applies only when `UTCI_SOURCE` is one of the in-script
computation paths (`compute_in_script_full` or `compute_in_script_package`),
which read raw ERA5 inputs. Under the recommended `cds_utci_product` path the
script instead downloads ECMWF's pre-computed daily UTCI directly (see the
next section), and this acquisition step is skipped.

When it runs, this step pulls the hourly inputs needed for UTCI: 2 m air
temperature (`2m_temperature`), 2 m dewpoint temperature
(`2m_dewpoint_temperature`), 10 m u-wind and v-wind components
(`10m_u_component_of_wind`, `10m_v_component_of_wind`), and surface downward
solar radiation (`surface_solar_radiation_downwards`). The full Di Napoli
MRT path (`compute_in_script_full`) additionally requires the full ERA5
single-levels radiation fields (`ssr`, `str`, `strd`, `fdir`), which are not
part of ERA5-Land.

# Obtain daily UTCI

The exposure source is the ECMWF pre-computed daily UTCI product — the same
product used in Chapter 2, read here at daily resolution. Each file is one
NetCDF per day (`ECMWF_utci_daily_stats_YYYYMMDD_v1.1_con*.nc`) containing
the layers `utci_daily_min` and `utci_daily_max`. Mean radiant temperature
is computed by ECMWF from the full ERA5 radiation fields, so the inflation
seen in an earlier in-script attempt — which approximated MRT from ERA5-Land
downward shortwave alone and produced daily-maximum UTCI near 35 °C in every
season — does not arise here. The grid is the ERA5 0.25° grid (about 28 km);
because the Kassena-Nankana catchment spans only a few grid cells, within-
catchment spatial variation in exposure is modest, a point acknowledged in
the chapter (§3.8.2, §3.10.3).

The exposure is the **daily maximum UTCI**, consistent with an acute-heat-
stress framing: the percentile thresholds and heatwave-run metrics below are
all defined on the same daily-maximum distribution, so the exposure
apparatus is internally coherent. The daily minimum is read only to allow
percentile diagnostics; no daily-mean exposure is constructed, and the
midrange of min and max is deliberately not used as a stand-in for the mean
(it is not the true 24-hour mean and its bias varies by season). The reader
extracts at the exact household coordinate — no displacement buffer is
applied, because the Navrongo geocodes are true household locations rather
than the jittered DHS cluster points used in Chapter 2.

For completeness the script retains alternative sources behind the
`UTCI_SOURCE` flag: a CDS download path for the same product, the full Di
Napoli (2018) in-script radiation balance (which requires full ERA5 inputs,
not ERA5-Land), and the deprecated ERA5-Land + simplified-MRT path retained
only to reproduce the first run.

## Recommended path — local ECMWF daily UTCI files

```{r local-daily, eval = RUN_LOCAL_DAILY}
# Read the ECMWF daily UTCI files already on disk in UTCI_DIR. For each unique
# (location, day) pair in the extraction grid, open the matching daily file
# and extract the daily MAXIMUM UTCI at the household coordinate. Daily max is
# the sole exposure; no daily-mean construct is built.

# 1. Discover and date-index the daily files.
daily_files <- list.files(UTCI_DIR, pattern = "\\.nc$", full.names = TRUE)
if (length(daily_files) == 0) {
  stop("No .nc files found in UTCI_DIR: ", UTCI_DIR)
}
file_index <- tibble(
  path = daily_files,
  day  = as.Date(vapply(daily_files, parse_date_from_filename, as.Date(NA)),
                 origin = "1970-01-01")
) |>
  filter(!is.na(day)) |>
  distinct(day, .keep_all = TRUE)

cat(sprintf("Daily UTCI files discovered: %s (dated %s to %s)\n",
            format(nrow(file_index), big.mark = ","),
            format(min(file_index$day)), format(max(file_index$day))))

# 2. Which calendar days are actually needed (from the extraction grid).
days_needed <- sort(unique(as.Date(extract_grid$day)))
files_to_read <- file_index |> filter(day %in% days_needed)

missing_days <- setdiff(as.character(days_needed), as.character(files_to_read$day))
if (length(missing_days) > 0) {
  warning(sprintf("%d needed days have no daily file (e.g. %s). These become NA.",
                  length(missing_days), paste(head(missing_days, 3), collapse = ", ")))
}
cat(sprintf("Days needed: %s; daily files available for them: %s\n",
            format(length(days_needed), big.mark = ","),
            format(nrow(files_to_read), big.mark = ",")))

# 3. Build the location point object once (true household coordinates).
loc_pts <- terra::vect(as.matrix(locations[, c("lon", "lat")]),
                       crs = "EPSG:4326")

# 4. Read each needed day once, extract daily-max UTCI at all locations.
read_one_day <- function(path, d) {
  # Repair zip-disguised NetCDFs if necessary.
  if (is_zipfile(path)) {
    fixed <- fix_zip_nc(path)
    if (is.na(fixed)) return(NULL)
    path <- fixed
  }
  r <- tryCatch(terra::rast(path), error = function(e) NULL)
  if (is.null(r)) return(NULL)

  r_max <- get_daily_layer(r, "max")
  v_max <- terra::extract(r_max, loc_pts)[, 2]

  # ECMWF UTCI ships in kelvin; convert to °C if needed.
  if (is_kelvin(r_max)) v_max <- v_max - 273.15

  tibble(loc_id = locations$loc_id, day = d, utci = v_max)
}

start_time <- Sys.time()
daily_list <- vector("list", nrow(files_to_read))
for (k in seq_len(nrow(files_to_read))) {
  daily_list[[k]] <- read_one_day(files_to_read$path[k], files_to_read$day[k])
  if (k %% 365L == 0L || k == nrow(files_to_read)) {
    elapsed <- difftime(Sys.time(), start_time, units = "mins")
    cat(sprintf("  %s / %s days read — %.1f min elapsed\n",
                format(k, big.mark = ","),
                format(nrow(files_to_read), big.mark = ","),
                as.numeric(elapsed)))
  }
}

daily_utci <- bind_rows(daily_list) |>
  distinct(loc_id, day, .keep_all = TRUE) |>
  mutate(day = as.Date(day))

saveRDS(daily_utci, file.path(LOG_DIR, "daily_utci_consolidated.rds"))
cat(sprintf(
  "\nDaily UTCI table assembled from local ECMWF files: %s rows (exposure = daily maximum UTCI)\n",
  format(nrow(daily_utci), big.mark = ",")
))
```

## Alternative path — download ECMWF daily UTCI from the CDS

```{r utci-product, eval = RUN_UTCI_PRODUCT}
# Self-guard for interactive runs (see compute-utci for rationale).
if (!isTRUE(RUN_UTCI_PRODUCT)) {
  message("Skipping CDS UTCI-product download/read (UTCI_SOURCE = '",
          UTCI_SOURCE, "').")
} else {

# Acquire and read ECMWF's pre-computed daily UTCI from the CDS
# "derived-utci-historical" dataset. The product provides UTCI at hourly
# resolution on the ERA5 grid (0.25°); we aggregate to daily max at each
# household coordinate.
#
# The product variable is "universal_thermal_climate_index"; values are in
# kelvin in the NetCDF and converted to °C here.

utci_product_request <- function(year, month) {
  first_day <- as.Date(sprintf("%04d-%02d-01", year, month))
  n_days    <- lubridate::days_in_month(first_day)
  list(
    dataset_short_name = "derived-utci-historical",
    variable           = "universal_thermal_climate_index",
    product_type       = "consolidated_dataset",
    version            = "1_1",
    year               = sprintf("%04d", year),
    month              = sprintf("%02d", month),
    day                = sprintf("%02d", seq_len(n_days)),
    area               = c(BBOX$lat_max, BBOX$lon_min,
                           BBOX$lat_min, BBOX$lon_max),
    data_format        = "netcdf",
    download_format    = "unarchived",
    target             = sprintf("utci_navrongo_%04d%02d.nc", year, month)
  )
}

# Which (year, month) tiles are needed, from the extraction grid.
extract_grid_ym <- extract_grid |>
  mutate(year = lubridate::year(day), month = lubridate::month(day))

month_index <- extract_grid_ym |>
  distinct(year, month) |>
  arrange(year, month) |>
  mutate(file = sprintf("utci_navrongo_%04d%02d.nc", year, month),
         path = file.path(UTCI_DIR, file))

# Download any tiles not already cached (only if acquisition is via CDS).
if (ERA5_ACQUISITION == "cds_download") {
  configure_cds()
  to_get <- month_index |> filter(!file.exists(path))
  if (nrow(to_get) > 0) {
    cat(sprintf("Downloading %d UTCI-product tiles from CDS...\n", nrow(to_get)))
    for (i in seq_len(nrow(to_get))) {
      req <- utci_product_request(to_get$year[i], to_get$month[i])
      ecmwfr::wf_request(request = req, transfer = TRUE,
                         path = UTCI_DIR, time_out = 3 * 3600, verbose = FALSE)
      if (i %% 12 == 0 || i == nrow(to_get)) {
        cat(sprintf("  %d / %d tiles downloaded\n", i, nrow(to_get)))
      }
    }
  } else {
    cat("All required UTCI-product tiles already cached.\n")
  }
}

# Read each tile, extract hourly UTCI at required household coordinates,
# aggregate to daily max and mean.
read_utci_product_tile <- function(nc_path, locations_pts, days_needed) {
  r <- terra::rast(nc_path)
  times <- parse_era5_times(r)                # reuse the shared time parser
  pts <- terra::vect(as.matrix(locations_pts[, c("lon", "lat")]),
                     crs = "EPSG:4326")
  vals <- terra::extract(r, pts)[, -1, drop = FALSE]
  vals <- as.matrix(vals)
  # Kelvin → °C if needed (UTCI product is distributed in K).
  if (mean(vals, na.rm = TRUE) > 100) vals <- vals - 273.15

  hour_dates  <- as.Date(times)
  unique_days <- sort(unique(hour_dates))
  if (!is.null(days_needed)) {
    unique_days <- sort(intersect(unique_days, as.Date(days_needed)))
  }

  map_dfr(unique_days, function(d) {
    cols <- which(hour_dates == d)
    if (length(cols) == 0) return(NULL)
    dmax  <- apply(vals[, cols, drop = FALSE], 1, max,  na.rm = TRUE)
    dmean <- rowMeans(vals[, cols, drop = FALSE], na.rm = TRUE)
    dmax[!is.finite(dmax)]   <- NA_real_
    dmean[!is.finite(dmean)] <- NA_real_
    tibble(loc_id = locations_pts$loc_id, day = d,
           utci = dmax, utci_mean = dmean)
  })
}

daily_list <- vector("list", nrow(month_index))
start_time <- Sys.time()
for (k in seq_len(nrow(month_index))) {
  yy <- month_index$year[k]; mm <- month_index$month[k]
  grid_month <- extract_grid_ym |> filter(year == yy, month == mm)
  locs_month <- locations |> filter(loc_id %in% unique(grid_month$loc_id))
  days_month <- unique(grid_month$day)

  daily_list[[k]] <- read_utci_product_tile(
    nc_path = month_index$path[k], locations_pts = locs_month,
    days_needed = days_month
  )
  if (k %% 12L == 0L || k == nrow(month_index)) {
    elapsed <- difftime(Sys.time(), start_time, units = "mins")
    cat(sprintf("  Tile %d / %d (%04d-%02d) — %.1f min elapsed\n",
                k, nrow(month_index), yy, mm, as.numeric(elapsed)))
  }
}

daily_utci <- bind_rows(daily_list) |>
  distinct(loc_id, day, .keep_all = TRUE) |>
  mutate(day = as.Date(day))

saveRDS(daily_utci, file.path(UTCI_DIR, "daily_utci_consolidated.rds"))
cat(sprintf(
  "\nDaily UTCI table written from pre-computed product: %s rows (exposure = daily maximum UTCI)\n",
  format(nrow(daily_utci), big.mark = ",")
))

}  # end interactive self-guard for utci-product
```

## In-script computation from ERA5-Land inputs (legacy / full-ERA5)

```{r compute-utci, eval = RUN_UTCI_COMPUTE}
# Self-guard: the knitr eval= flag skips this chunk during rendering, but it
# does NOT protect against running the document interactively (line by line in
# the console), where eval= has no effect. This guard makes the chunk exit
# cleanly in both cases, so the default local_ecmwf_daily run is never
# interrupted by the legacy in-script ERA5-Land path.
if (!isTRUE(RUN_UTCI_COMPUTE)) {
  message("Skipping in-script ERA5-Land UTCI computation (UTCI_SOURCE = '",
          UTCI_SOURCE, "').")
} else {

# For each (year, month) tile needed by extract_grid, read ERA5-Land inputs,
# compute hourly UTCI only at required household coordinates and calendar
# days, aggregate to daily max and daily mean, and write a consolidated table.

# ----------------------------------------------------------------------------
# Helper: De-accumulate ERA5-Land hourly accumulated radiation.
# ERA5-Land surface_solar_radiation_downwards (SSRD) is accumulated from
# 00:00 UTC to the forecast hour. To recover the per-hour mean W/m² for
# each hour-ending timestamp, we take the difference between successive
# accumulated values and divide by 3600 (seconds in an hour). The 01:00
# value is the accumulation from 00:00 to 01:00 and is used directly.
# A daily reset is assumed at every 00:00 UTC timestamp.
# ----------------------------------------------------------------------------
deaccumulate_ssrd <- function(ssrd_accum_matrix, hour_of_day) {
  # ssrd_accum_matrix: locations × hours, in J/m² (accumulated since 00 UTC)
  # hour_of_day: integer 0..23 corresponding to each column of the matrix
  n_loc <- nrow(ssrd_accum_matrix)
  n_t   <- ncol(ssrd_accum_matrix)
  out   <- matrix(NA_real_, n_loc, n_t)
  for (j in seq_len(n_t)) {
    if (hour_of_day[j] == 0) {
      # 00:00 timestamp: accumulation has just reset; no preceding hour.
      out[, j] <- 0
    } else if (j == 1 || hour_of_day[j] == 1) {
      # 01:00 timestamp: accumulation from 00:00 to 01:00, use directly.
      out[, j] <- ssrd_accum_matrix[, j] / 3600
    } else {
      # All other hours: difference from previous hour, divide by 3600.
      out[, j] <- (ssrd_accum_matrix[, j] - ssrd_accum_matrix[, j - 1]) / 3600
    }
  }
  # SSRD must be non-negative; clamp small negative numerical artefacts to 0.
  out[out < 0] <- 0
  out
}

# ----------------------------------------------------------------------------
# Helper: Magnus formula for saturation vapour pressure (hPa) from T (°C).
# ----------------------------------------------------------------------------
e_sat <- function(T_C) 6.112 * exp((17.62 * T_C) / (243.12 + T_C))

# ----------------------------------------------------------------------------
# Parse the hourly time axis from an ERA5-Land SpatRaster.
# CDS NetCDFs often encode valid_time in layer names (e.g. t2m_valid_time=...)
# rather than a GDAL-readable time dimension, so terra::time() returns NA.
# ----------------------------------------------------------------------------
parse_era5_times <- function(spat_rast) {
  t <- terra::time(spat_rast)
  if (length(t) == terra::nlyr(spat_rast) && !all(is.na(t))) {
    return(t)
  }
  nm <- names(spat_rast)
  if (any(grepl("valid_time=", nm, fixed = TRUE))) {
    unix_sec <- as.numeric(sub(".*valid_time=", "", nm))
    return(as.POSIXct(unix_sec, origin = "1970-01-01", tz = "UTC"))
  }
  stop(sprintf("Could not determine time axis for %s", basename(terra::sources(spat_rast)[1])))
}

# ----------------------------------------------------------------------------
# Read one ERA5-Land variable from a NetCDF, returning a matrix of
# (locations × time) values with associated timestamps. Handles both the
# short-name and long-name variable conventions that CDS may return.
# Optional time_indices restricts extraction to a subset of hourly layers.
# ----------------------------------------------------------------------------
read_era5_variable <- function(nc_path, var_candidates, locations_pts,
                               time_indices = NULL) {
  for (vn in var_candidates) {
    sub_r <- tryCatch(
      terra::rast(nc_path, subds = vn),
      error = function(e) NULL
    )
    if (!is.null(sub_r)) {
      times <- parse_era5_times(sub_r)
      if (!is.null(time_indices)) {
        sub_r <- sub_r[[time_indices]]
        times <- times[time_indices]
      }
      pts <- terra::vect(as.matrix(locations_pts[, c("lon", "lat")]),
                         crs = "EPSG:4326")
      ex  <- terra::extract(sub_r, pts)[, -1, drop = FALSE]
      return(list(
        values    = as.matrix(ex),
        times     = times,
        var_used  = vn
      ))
    }
  }
  stop(sprintf("None of the candidate variable names found in %s: %s",
               basename(nc_path), paste(var_candidates, collapse = ", ")))
}

# ----------------------------------------------------------------------------
# Resolve hourly layer indices for the calendar days required from a tile.
# ----------------------------------------------------------------------------
era5_time_indices <- function(nc_path, var_candidates, days_needed) {
  days_needed <- as.Date(days_needed)
  for (vn in var_candidates) {
    sub_r <- tryCatch(
      terra::rast(nc_path, subds = vn),
      error = function(e) NULL
    )
    if (!is.null(sub_r)) {
      hour_dates <- as.Date(parse_era5_times(sub_r))
      idx <- which(hour_dates %in% days_needed)
      return(idx)
    }
  }
  stop(sprintf("Could not resolve time axis in %s", basename(nc_path)))
}

# ----------------------------------------------------------------------------
# Vectorized UTCI via comf's Bröde polynomial.
# comf::calcUTCI() accepts scalars only; utciApprox() is the same polynomial
# and accepts vector inputs (required for location × hour matrices).
# ----------------------------------------------------------------------------
calc_utci_vec <- function(ta, tr, vel, rh) {
  utci_approx <- getFromNamespace("utciApprox", "comf")
  utci_approx(
    ta  = pmin(pmax(ta, 0), 40),
    tr  = pmin(pmax(tr, 0), 40),
    vel = pmin(pmax(vel, 0), 10),
    rh  = pmin(pmax(rh, 0), 100)
  )
}

# ----------------------------------------------------------------------------
# Main: compute daily UTCI summaries for one ERA5-Land month-tile.
# Returns daily max (utci) and daily mean (utci_mean) from the same hourly
# series. Restrict to days_needed and locations_pts for efficiency.
# ----------------------------------------------------------------------------
compute_daily_utci <- function(nc_path, locations_pts, days_needed = NULL) {
  if (nrow(locations_pts) == 0) {
    return(tibble(
      loc_id    = integer(),
      day       = as.Date(character()),
      utci      = double(),
      utci_mean = double()
    ))
  }

  time_indices <- if (!is.null(days_needed)) {
    idx <- era5_time_indices(nc_path, c("t2m", "2m_temperature"), days_needed)
    if (length(idx) == 0) return(tibble(
      loc_id = integer(), day = as.Date(character()),
      utci = double(), utci_mean = double()
    ))
    idx
  } else {
    NULL
  }

  t2m   <- read_era5_variable(nc_path,
              c("t2m", "2m_temperature"), locations_pts, time_indices)
  d2m   <- read_era5_variable(nc_path,
              c("d2m", "2m_dewpoint_temperature"), locations_pts, time_indices)
  u10   <- read_era5_variable(nc_path,
              c("u10", "10m_u_component_of_wind"), locations_pts, time_indices)
  v10   <- read_era5_variable(nc_path,
              c("v10", "10m_v_component_of_wind"), locations_pts, time_indices)
  ssrd  <- read_era5_variable(nc_path,
              c("ssrd", "surface_solar_radiation_downwards"), locations_pts, time_indices)

  stopifnot(length(t2m$times) == length(d2m$times),
            length(t2m$times) == length(u10$times),
            length(t2m$times) == length(v10$times),
            length(t2m$times) == length(ssrd$times))

  t2m_C   <- t2m$values - 273.15
  d2m_C   <- d2m$values - 273.15
  wind_ms <- sqrt(u10$values^2 + v10$values^2)

  hour_of_day <- lubridate::hour(t2m$times)
  ssrd_Wm2    <- deaccumulate_ssrd(ssrd$values, hour_of_day)

  rh_pct <- 100 * e_sat(d2m_C) / e_sat(t2m_C)
  rh_pct[rh_pct > 100] <- 100

  # --------------------------------------------------------------------------
  # Mean radiant temperature.
  # --------------------------------------------------------------------------
  if (UTCI_SOURCE == "compute_in_script_full") {
    # Full Di Napoli et al. (2018) radiation balance. This REQUIRES the full
    # ERA5 single-levels radiation fields, which must have been read alongside
    # the four basic inputs above:
    #   ssr  — surface net solar radiation (J/m², accumulated)
    #   str  — surface net thermal radiation (J/m², accumulated)
    #   strd — surface thermal radiation downwards (J/m², accumulated)
    #   fdir — total-sky direct solar radiation at surface (J/m², accumulated)
    # plus the solar zenith angle, computed from latitude/longitude/time.
    #
    # MRT is the temperature of a uniform enclosure giving the same radiant
    # load as the actual environment:
    #   S_str = MRT_long + (alpha_ir / eps_p) * MRT_short
    # where the longwave term combines downward and upward longwave, and the
    # shortwave term combines diffuse, reflected, and (zenith-weighted) direct
    # beam. fp(zenith) is the projected-area factor for a standing person.
    needed_full <- c("ssr", "str", "strd", "fdir")
    have_full   <- exists("rad_full") && all(needed_full %in% names(rad_full))
    if (!have_full) {
      stop("UTCI_SOURCE = 'compute_in_script_full' requires the full ERA5 ",
           "radiation fields (ssr, str, strd, fdir) to be read into `rad_full`. ",
           "These are NOT available in ERA5-Land. Stage full ERA5 single-levels ",
           "inputs, or use UTCI_SOURCE = 'cds_utci_product' instead.")
    }

    sigma  <- 5.67e-8
    alpha_ir <- 0.70   # shortwave absorption coefficient of the human body
    eps_p    <- 0.97   # emissivity / projected-area emission factor
    # De-accumulate the radiation fields to hourly mean W/m².
    ssr_W  <- deaccumulate_ssrd(rad_full$ssr,  hour_of_day)
    str_W  <- deaccumulate_ssrd(rad_full$str,  hour_of_day)
    strd_W <- deaccumulate_ssrd(rad_full$strd, hour_of_day)
    fdir_W <- deaccumulate_ssrd(rad_full$fdir, hour_of_day)

    # Solar zenith angle (radians) at each (location, time).
    cossza <- solar_cos_zenith(locations_pts$lat, locations_pts$lon, t2m$times)
    cossza[cossza < 0] <- 0  # sun below horizon → no direct beam

    # Projected-area factor for a standing/walking person (Di Napoli 2018).
    fp <- 0.308 * cos((pi/180) * (acos(cossza) * 180/pi) * (0.998 -
            ((acos(cossza) * 180/pi)^2) / 50000))

    # Diffuse shortwave = downward shortwave − direct beam.
    # Downward shortwave is recovered as ssr_down = ssr_net / (1 - albedo_sfc);
    # ERA5 provides net, so use the standard surface albedo for partition.
    albedo_sfc <- 0.30
    sw_down    <- ssrd_Wm2                # already the downward shortwave
    sw_diffuse <- pmax(sw_down - fdir_W, 0)
    sw_reflect <- albedo_sfc * sw_down

    # Mean radiant flux density (W/m²): hemispheric longwave (down + up) plus
    # absorbed shortwave (diffuse + reflected isotropic, direct via fp).
    lw_up   <- strd_W - str_W             # upward longwave = down − net
    Rlong   <- 0.5 * strd_W + 0.5 * lw_up
    Rshort  <- (alpha_ir / eps_p) *
               (0.5 * sw_diffuse + 0.5 * sw_reflect + fp * fdir_W)
    mrt_K   <- ((Rlong + Rshort) / sigma)^0.25
    mrt_C   <- mrt_K - 273.15

  } else {
    # DEPRECATED simplified MRT (UTCI_SOURCE = "compute_in_script_package").
    # Retained only to reproduce the first run. Known to over-weight daytime
    # shortwave and inflate daily-maximum UTCI; not for analytic use. See the
    # narrative above and prefer UTCI_SOURCE = "cds_utci_product".
    sigma  <- 5.67e-8
    albedo <- 0.30
    eps_p  <- 0.97
    mrt_K  <- ((ssrd_Wm2 * (1 - albedo) / (sigma * eps_p)) +
               (t2m_C + 273.15)^4)^0.25
    mrt_C  <- mrt_K - 273.15
  }

  n_loc <- nrow(t2m_C)
  n_t   <- ncol(t2m_C)
  utci_hourly <- matrix(NA_real_, n_loc, n_t)

  if (UTCI_SOURCE %in% c("compute_in_script_full", "compute_in_script_package")) {
    utci_hourly <- matrix(
      calc_utci_vec(
        ta  = as.vector(t2m_C),
        tr  = as.vector(mrt_C),
        vel = as.vector(wind_ms),
        rh  = as.vector(rh_pct)
      ),
      nrow = n_loc,
      ncol = n_t
    )
  } else {
    if (file.exists("utci_brode.R")) {
      source("utci_brode.R")
      utci_hourly <- utci_brode(t2m_C, mrt_C, wind_ms, rh_pct)
    } else {
      stop("UTCI_SOURCE = 'compute_in_script_manual' but utci_brode.R is not present.")
    }
  }

  hour_dates  <- as.Date(t2m$times)
  unique_days <- sort(unique(hour_dates))
  if (!is.null(days_needed)) {
    unique_days <- sort(intersect(unique_days, as.Date(days_needed)))
  }

  daily <- map_dfr(unique_days, function(d) {
    cols_today <- which(hour_dates == d)
    if (length(cols_today) == 0) return(NULL)
    utci_max <- apply(utci_hourly[, cols_today, drop = FALSE], 1,
                      max, na.rm = TRUE)
    utci_avg <- rowMeans(utci_hourly[, cols_today, drop = FALSE], na.rm = TRUE)
    utci_max[!is.finite(utci_max)] <- NA_real_
    utci_avg[!is.finite(utci_avg)] <- NA_real_
    tibble(
      loc_id    = locations_pts$loc_id,
      day       = as.Date(d),
      utci      = utci_max,
      utci_mean = utci_avg
    )
  })

  daily
}

# Only process month-tiles and (location, day) pairs present in extract_grid.
extract_grid_ym <- extract_grid |>
  mutate(year = lubridate::year(day), month = lubridate::month(day))

months_needed <- extract_grid_ym |>
  distinct(year, month) |>
  arrange(year, month)

month_files_all <- list.files(
  CLIMATE_DIR,
  pattern = "^era5_land_navrongo_\\d{6}\\.nc$",
  full.names = TRUE
)

parse_month_file <- function(path) {
  ym <- sub(".*era5_land_navrongo_(\\d{6})\\.nc$", "\\1", basename(path))
  tibble(
    path  = path,
    year  = as.integer(substr(ym, 1L, 4L)),
    month = as.integer(substr(ym, 5L, 6L))
  )
}

month_index <- month_files_all |>
  parse_month_file() |>
  inner_join(months_needed, by = c("year", "month")) |>
  arrange(year, month)

cat(sprintf(
  "UTCI computation scoped to %d month-tiles (of %d cached) covering %s (loc, day) pairs.\n",
  nrow(month_index),
  length(month_files_all),
  format(nrow(extract_grid), big.mark = ",")
))

daily_utci_list <- vector("list", nrow(month_index))
start_time <- Sys.time()

for (k in seq_len(nrow(month_index))) {
  yy <- month_index$year[k]
  mm <- month_index$month[k]
  grid_month <- extract_grid_ym |>
    filter(year == yy, month == mm)
  locs_month <- locations |>
    filter(loc_id %in% unique(grid_month$loc_id))
  days_month <- unique(grid_month$day)

  daily_utci_list[[k]] <- compute_daily_utci(
    nc_path        = month_index$path[k],
    locations_pts  = locs_month,
    days_needed    = days_month
  )

  if (k %% 12L == 0L || k == nrow(month_index)) {
    elapsed <- difftime(Sys.time(), start_time, units = "mins")
    cat(sprintf("  Tile %d / %d (%04d-%02d) — %.1f min elapsed\n",
                k, nrow(month_index), yy, mm, as.numeric(elapsed)))
  }
}

daily_utci <- bind_rows(daily_utci_list) |>
  distinct(loc_id, day, .keep_all = TRUE) |>
  mutate(day = as.Date(day))

saveRDS(daily_utci, file.path(UTCI_DIR, "daily_utci_consolidated.rds"))
cat(sprintf(
  "\nDaily UTCI table written: %s rows (exposure = daily maximum UTCI)\n",
  format(nrow(daily_utci), big.mark = ",")
))

}  # end interactive self-guard for compute-utci
```


# Assemble pregnancy-level exposure matrix

For each pregnancy in the analytic cohort, we now construct the columns
that the downstream DLNM, case-crossover, and trimester-regression scripts
consume.

```{r assemble-exposure}
# Join daily UTCI values (daily maximum — the sole exposure) to the
# (pregnancy, day) table.
preg_daily <- preg_days |>
  mutate(day = as.Date(day)) |>
  left_join(
    daily_utci |>
      dplyr::select(loc_id, day, utci) |>
      mutate(day = as.Date(day)),
    by = c("loc_id", "day")
  )

n_total  <- nrow(preg_daily)
n_filled <- sum(!is.na(preg_daily$utci))
cat(sprintf("Pregnancy-day cells filled (daily max UTCI): %s of %s (%.2f%%)\n",
            format(n_filled, big.mark = ","),
            format(n_total,  big.mark = ","),
            100 * n_filled / n_total))

# Shared helper: window summaries from a chosen daily UTCI column.
window_summaries <- function(preg_daily_df, utci_col, suffix = "") {
  preg_with_dates <- preg_daily_df |>
    inner_join(preg |> dplyr::select(row_id, child_dob, preg_start_assumed),
               by = "row_id") |>
    mutate(days_from_conception = as.integer(day - preg_start_assumed))

  trimester <- preg_with_dates |>
    mutate(trimester = case_when(
      days_from_conception >= 0L & days_from_conception <= TRIMESTER1_END ~ "t1",
      days_from_conception > TRIMESTER1_END &
        days_from_conception <= TRIMESTER2_END ~ "t2",
      days_from_conception > TRIMESTER2_END & day <= child_dob ~ "t3",
      TRUE ~ NA_character_
    )) |>
    filter(!is.na(trimester)) |>
    group_by(row_id, trimester) |>
    summarise(
      mean_val = if (all(is.na(.data[[utci_col]]))) {
        NA_real_
      } else {
        mean(.data[[utci_col]], na.rm = TRUE)
      },
      .groups = "drop"
    ) |>
    pivot_wider(names_from = trimester, values_from = mean_val, names_prefix = "") |>
    rename(
      !!paste0("t1_meanmax_utci", suffix) := t1,
      !!paste0("t2_meanmax_utci", suffix) := t2,
      !!paste0("t3_meanmax_utci", suffix) := t3
    )

  preg_mean <- preg_with_dates |>
    filter(day >= preg_start_assumed, day <= child_dob) |>
    group_by(row_id) |>
    summarise(
      !!paste0("preg_meanmax_utci", suffix) := if (all(is.na(.data[[utci_col]]))) {
        NA_real_
      } else {
        mean(.data[[utci_col]], na.rm = TRUE)
      },
      .groups = "drop"
    )

  short_windows <- reduce(
    map(OUTCOME_WINDOWS, function(n) {
      preg_with_dates |>
        mutate(lag = as.integer(child_dob - day)) |>
        filter(lag >= 0L, lag < n) |>
        group_by(row_id) |>
        summarise(
          !!paste0("last_", n, "d_meanmax_utci", suffix) := if (all(is.na(.data[[utci_col]]))) {
            NA_real_
          } else {
            mean(.data[[utci_col]], na.rm = TRUE)
          },
          .groups = "drop"
        )
    }),
    function(a, b) full_join(a, b, by = "row_id")
  )

  list(trimester = trimester, preg_mean = preg_mean, short_windows = short_windows)
}

# Window summaries are means of the DAILY-MAXIMUM UTCI over each window
# (i.e. the average peak heat stress across the window), not daily means.
primary_summaries <- window_summaries(preg_daily, "utci", suffix = "")

trimester_means <- primary_summaries$trimester
preg_mean       <- primary_summaries$preg_mean
short_windows   <- primary_summaries$short_windows

# ----------------------------------------------------------------------------
# DLNM daily lag matrix: one row per pregnancy, columns lag_0 ... lag_LAG_MAX.
# Uses daily-max UTCI (primary statistic).
# ----------------------------------------------------------------------------
lag_matrix <- preg_daily |>
  inner_join(preg |> dplyr::select(row_id, child_dob), by = "row_id") |>
  mutate(lag = as.integer(child_dob - day)) |>
  filter(lag >= LAG_MIN, lag <= LAG_MAX) |>
  pivot_wider(id_cols     = row_id,
              names_from  = lag,
              values_from = utci,
              names_prefix = "lag_")

lag_cols <- paste0("lag_", LAG_MIN:LAG_MAX)
missing_lag_cols <- setdiff(lag_cols, names(lag_matrix))
for (m in missing_lag_cols) lag_matrix[[m]] <- NA_real_
lag_matrix <- lag_matrix[, c("row_id", lag_cols)]

coverage <- lag_matrix |>
  mutate(lag_cells_filled = rowSums(!is.na(across(starts_with("lag_")))),
         lag_coverage_pct = round(100 * lag_cells_filled / (LAG_MAX - LAG_MIN + 1L), 1)) |>
  dplyr::select(row_id, lag_cells_filled, lag_coverage_pct)

exposure <- preg |>
  dplyr::select(row_id, child_id, motherid, locationid, lon, lat,
                child_dob, preg_start_assumed) |>
  left_join(lag_matrix, by = "row_id") |>
  left_join(trimester_means, by = "row_id") |>
  left_join(preg_mean, by = "row_id") |>
  left_join(short_windows, by = "row_id") |>
  left_join(coverage, by = "row_id")
```

# Percentile and heatwave exposure metrics

These metrics, carried over from the Chapter 2 daily pipeline, summarise heat
*intensity* and *persistence* over each pregnancy window, complementing the
mean-based summaries above and the daily lag matrix. Percentile thresholds
are **location-specific**: each grid cell's own daily-max UTCI distribution
over the analytic window defines its p90 and p95. For each pregnancy and each
window (full pregnancy and the three trimesters) the script computes the
number of days at or above each threshold, the cumulative degree-days of
exceedance above the threshold, and the longest consecutive run at or above
the p90 threshold (a heatwave-duration measure).

```{r heat-metrics}
# 1. Location-specific percentile thresholds from each cell's daily-max series
#    over the full analytic window (using the consolidated daily_utci table).
loc_thresholds <- daily_utci |>
  group_by(loc_id) |>
  summarise(
    thr_p90 = quantile(utci, PERCENTILES[1] / 100, na.rm = TRUE),
    thr_p95 = quantile(utci, PERCENTILES[2] / 100, na.rm = TRUE),
    .groups = "drop"
  )

# 2. Tag each pregnancy-day with its window membership and exceedance flags.
preg_day_tagged <- preg_daily |>
  # preg_daily already carries loc_id (from preg_days), so import only the
  # date columns from preg here — re-importing loc_id would create loc_id.x /
  # loc_id.y and break the loc_thresholds join below.
  inner_join(preg |> dplyr::select(row_id, child_dob, preg_start_assumed),
             by = "row_id") |>
  left_join(loc_thresholds, by = "loc_id") |>
  mutate(
    days_from_conception = as.integer(day - preg_start_assumed),
    window = case_when(
      day < preg_start_assumed | day > child_dob ~ NA_character_,
      days_from_conception <= TRIMESTER1_END                              ~ "t1",
      days_from_conception <= TRIMESTER2_END                              ~ "t2",
      TRUE                                                                ~ "t3"
    ),
    exc90 = as.integer(!is.na(utci) & utci >= thr_p90),
    exc95 = as.integer(!is.na(utci) & utci >= thr_p95),
    cumexc90 = pmax(utci - thr_p90, 0),
    cumexc95 = pmax(utci - thr_p95, 0)
  )

# 3. Window-level day-count and cumulative-exceedance metrics.
window_metrics <- function(df, win_label, suffix) {
  d <- if (win_label == "preg") {
    df |> filter(!is.na(window))
  } else {
    df |> filter(window == win_label)
  }
  d |>
    group_by(row_id) |>
    summarise(
      "{suffix}_p90_days"   := sum(exc90, na.rm = TRUE),
      "{suffix}_p95_days"   := sum(exc95, na.rm = TRUE),
      "{suffix}_p90_cumexc" := sum(cumexc90, na.rm = TRUE),
      "{suffix}_p95_cumexc" := sum(cumexc95, na.rm = TRUE),
      "{suffix}_ndays"      := sum(!is.na(utci)),
      .groups = "drop"
    )
}

# 4. Longest consecutive run at or above p90 within a window (heatwave length).
max_run_metric <- function(df, win_label, suffix) {
  d <- if (win_label == "preg") {
    df |> filter(!is.na(window))
  } else {
    df |> filter(window == win_label)
  }
  d |>
    arrange(row_id, day) |>
    group_by(row_id) |>
    summarise(
      "{suffix}_p90_maxrun" := {
        x <- exc90[order(day)]
        x[is.na(x)] <- 0L
        if (length(x) == 0) 0L else max(rle(x)$lengths[rle(x)$values == 1L], 0L)
      },
      .groups = "drop"
    )
}

heat_metrics <- preg |> dplyr::select(row_id)
for (w in c("preg", "t1", "t2", "t3")) {
  heat_metrics <- heat_metrics |>
    left_join(window_metrics(preg_day_tagged, w, w), by = "row_id") |>
    left_join(max_run_metric(preg_day_tagged, w, w), by = "row_id")
}

# Attach to the main exposure table and record the thresholds used.
exposure <- exposure |>
  left_join(heat_metrics, by = "row_id") |>
  left_join(preg |> dplyr::select(row_id, loc_id), by = "row_id") |>
  left_join(loc_thresholds, by = "loc_id") |>
  dplyr::select(-loc_id)

cat(sprintf("Heat metrics computed for %s pregnancies across %s grid cells.\n",
            format(nrow(exposure), big.mark = ","),
            format(nrow(loc_thresholds), big.mark = ",")))
```

# Diagnostics

```{r diag-tables}
# Window summaries are means of the daily-MAXIMUM UTCI over each window.
exposure_vars <- c(
  "t1_meanmax_utci", "t2_meanmax_utci", "t3_meanmax_utci", "preg_meanmax_utci",
  paste0("last_", OUTCOME_WINDOWS, "d_meanmax_utci")
)

# Coverage summary
coverage_summary <- tibble(
  variable = exposure_vars,
  pct_non_missing = sapply(exposure_vars, function(v)
    round(100 * mean(!is.na(exposure[[v]])), 1)),
  median_utci = sapply(exposure_vars, function(v)
    round(median(exposure[[v]], na.rm = TRUE), 1)),
  p10 = sapply(exposure_vars, function(v)
    round(quantile(exposure[[v]], 0.10, na.rm = TRUE), 1)),
  p90 = sapply(exposure_vars, function(v)
    round(quantile(exposure[[v]], 0.90, na.rm = TRUE), 1))
)

nice_table(coverage_summary,
           caption = "Per-pregnancy exposure summary — mean of daily-maximum UTCI over each window")

# DLNM lag-matrix coverage
lag_coverage_summary <- tibble(
  metric = c("pregnancies with 100% lag coverage",
             "pregnancies with ≥95% lag coverage",
             "pregnancies with ≥80% lag coverage",
             "median lag coverage (%)"),
  value  = c(sum(exposure$lag_coverage_pct == 100),
             sum(exposure$lag_coverage_pct >= 95),
             sum(exposure$lag_coverage_pct >= 80),
             round(median(exposure$lag_coverage_pct), 1))
)
nice_table(lag_coverage_summary,
           caption = "DLNM lag-matrix coverage (lag_0 through lag_30)")

# Heat-metric summary (carried over from Chapter 2)
heat_summary <- tibble(
  metric = c("median p90 threshold (°C, across cells)",
             "median p95 threshold (°C, across cells)",
             "median pregnancy days ≥ p90",
             "median pregnancy days ≥ p95",
             "median longest p90 run in pregnancy (days)",
             "median T3 days ≥ p90"),
  value  = c(round(median(exposure$thr_p90, na.rm = TRUE), 1),
             round(median(exposure$thr_p95, na.rm = TRUE), 1),
             round(median(exposure$preg_p90_days, na.rm = TRUE), 1),
             round(median(exposure$preg_p95_days, na.rm = TRUE), 1),
             round(median(exposure$preg_p90_maxrun, na.rm = TRUE), 1),
             round(median(exposure$t3_p90_days, na.rm = TRUE), 1))
)
nice_table(heat_summary,
           caption = "Heat-intensity and heatwave metrics (location-specific percentiles)")
```

# Persist outputs

```{r save}
saveRDS(exposure, file.path(OUT_DIR, "exposure.rds"))
write_csv(exposure |> dplyr::select(-starts_with("lag_")),
          file.path(OUT_DIR, "exposure_summaries.csv"))

cat(sprintf("Files written:\n  %s\n  %s\n",
            file.path(OUT_DIR, "exposure.rds"),
            file.path(OUT_DIR, "exposure_summaries.csv")))
```

# Expected diagnostic counts

The script has been developed against the development sample produced by
`01_build_spine.Rmd`. Sample-stage and full-data expected counts are
documented below; the full-data figures will be reported once the script
is executed against the full extract by NHRC analysts.

```{r expected, echo = FALSE}
tibble(
  metric = c("pregnancies entering exposure linkage (full extract)",
             "unique household coordinates (full extract)",
             "(location, day) pairs to extract",
             "ERA5-Land month-tiles required",
             "expected (pregnancy, day) cells in lag matrix",
             "expected (pregnancy, day) cells in pregnancy window"),
  expected_value = c("approximately 94,649 (97.5% of 97,045)",
                     "approximately 4,000–6,000",
                     "approximately 23,500,000",
                     "312 (1995–2020 × 12 months)",
                     paste0("approximately ", format(94649 * 31, big.mark = ",")),
                     paste0("approximately ", format(94649 * 280, big.mark = ",")))
) |>
  nice_table(caption = "Expected diagnostic counts at full-data execution",
             wrap_col = 1, wrap_width = "9cm")
```

# Session information

```{r session-info}
sessionInfo()
```

# Next steps

With `exposure.rds` in hand, the immediate items are, in order:

1. **Validate the exposure construction** against substantive checks: the
   seasonal pattern of daily UTCI within the catchment (peaks in
   hot-dry season, troughs in dry-cool harmattan), the spatial pattern
   across the catchment (which should be modest given the small
   geographic extent), and the year-on-year trend (which should track
   the regional warming signal). These checks belong in
   `03_phase1_descriptive.Rmd` rather than this script.

2. **Construct the DLNM cross-basis** in `05_phase3_dlnm_perinatal.Rmd`.
   The cross-basis takes the daily lag matrix from `exposure.rds` and
   produces a set of model-matrix columns via `dlnm::crossbasis()`. The
   cross-basis is the substantive output that enters the Bayesian DLNM in
   Phase 3.

3. **Run the community-centroid sensitivity branch** in
   `09_phase7_sensitivity.Rmd`. For the 2.5% of pregnancies without
   household coordinates, the climate linkage uses a community-centroid
   coordinate derived from `locationid`-level aggregation of the geocoded
   subset. The structure of `02_link_era5_utci.Rmd` is reusable for this
   branch with a single change at the GPS-coverage gate.
