---
title: "Linking Daily UTCI Exposure to Pregnancies — Navrongo HDSS"
subtitle: "Heat Stress and Perinatal Outcomes in Ghana"
author: "Vincent"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
    df_print: paged
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  comment = "#>"
)
```

# Purpose and scope

This document implements the climate-linkage step of the analytic plan
accompanying Chapter 3 (Methodology, Version 8.0). It consumes the
pregnancy-level spine produced by `01_build_spine.Rmd` and produces a
pregnancy-level **daily UTCI exposure matrix** for the 0–30 day window
pre-delivery, together with trimester and outcome-anchored summaries that
feed downstream DLNM, case-crossover, and sensitivity analyses.

Three operations are executed here.

- **(i) ERA5-Land input acquisition.** Four hourly variables (2 m air
  temperature, 2 m dewpoint temperature, 10 m wind speed, downward
  shortwave radiation) are obtained for the Navrongo catchment bounding box
  across the 1995–2020 analytic window. The acquisition method is
  configurable: download-at-runtime via the Copernicus CDS API, or
  read-from-disk if the inputs are already staged.
- **(ii) Daily UTCI computation.** Hourly UTCI is computed at each unique
  household coordinate using the Bröde et al. (2012) polynomial
  approximation, then aggregated to daily summaries (daily maximum as the
  primary statistic, daily mean as the sensitivity). The UTCI calculation
  follows Di Napoli et al. (2018) for the mean radiant temperature
  approximation suitable for gridded reanalysis inputs. The computation
  method is configurable to support either in-script computation or
  reading from a pre-computed daily UTCI archive.
- **(iii) Pregnancy-level exposure assembly.** For each pregnancy in the
  spine, the script extracts the daily UTCI series for the 0–30 day window
  pre-delivery (the DLNM lag matrix), the trimester means (T1, T2, T3,
  full-pregnancy), and the outcome-anchored short-window means (last 7,
  14, 30 days). The resulting pregnancy-level exposure matrix is written
  to `exposure.rds` for downstream DLNM and case-crossover scripts.

**Note on workflow.** Per the code-then-execute arrangement agreed with
NHRC, the researcher develops and tests this script against a sample
dataset and stages the inputs for NHRC analysts to execute against the
full extract on the Centre's secure infrastructure. The configuration
block below exposes all environment-dependent paths and credentials so
that the script runs identically on the development environment and on
the Centre's compute environment.

# Configuration

The configuration block below exposes the operational choices flagged in
the chapter discussion of `02_link_era5_utci.Rmd`. Three of these reflect
unresolved decisions that may be settled in conversation with NHRC; the
script is written to support each option without restructuring.

```{r config}
# ============================================================================
# Paths
# ============================================================================
IN_PATH      <- "spine.rds"                # output of 01_build_spine.Rmd
OUT_DIR      <- "."                         # where exposure.rds is written
CLIMATE_DIR  <- "era5_land_navrongo"        # local cache of ERA5-Land inputs
UTCI_DIR     <- file.path(CLIMATE_DIR, "utci_daily")  # daily UTCI rasters
LOG_DIR      <- file.path(OUT_DIR, "logs")  # extraction diagnostics

dir.create(CLIMATE_DIR, showWarnings = FALSE, recursive = TRUE)
dir.create(UTCI_DIR,    showWarnings = FALSE, recursive = TRUE)
dir.create(LOG_DIR,     showWarnings = FALSE, recursive = TRUE)

# ============================================================================
# Catchment bounding box (Kassena-Nankana districts, Upper East Region)
# Slight padding on each side to ensure no household falls on a tile edge.
# ============================================================================
BBOX <- list(
  lat_min = 10.50, lat_max = 11.20,
  lon_min = -1.45, lon_max = -0.75
)

# ============================================================================
# Analytic window (matches Chapter 3 §3.8.1)
# ============================================================================
YEAR_MIN <- 1995L
YEAR_MAX <- 2020L

# ============================================================================
# Exposure construction parameters (matches Chapter 3 §3.10.3)
# ============================================================================
LAG_MIN          <- 0L    # day of delivery
LAG_MAX          <- 30L   # 30 days before delivery
GESTATION_DAYS   <- 280L  # term gestation; sensitivity at 259 and 294
TRIMESTER1_END   <- 91L   # days 0–91 post-conception
TRIMESTER2_END   <- 182L  # days 92–182
# Trimester 3 = days 183–280 (delivery)
OUTCOME_WINDOWS  <- c(7L, 14L, 30L)  # last-N-day means

# ============================================================================
# Daily UTCI summary statistic (primary vs sensitivity)
# ============================================================================
DAILY_STAT_PRIMARY     <- "max"   # daily maximum UTCI
DAILY_STAT_SENSITIVITY <- "mean"  # daily mean UTCI

# ============================================================================
# OPERATIONAL CHOICES — to be confirmed with NHRC climate-expertise lead
# ============================================================================

UTCI_SOURCE <- "compute_in_script_package"

ERA5_ACQUISITION <- "cds_download"

# ============================================================================
# Copernicus CDS credentials (used only if ERA5_ACQUISITION == "cds_download")
#
# Before running this script on a new machine, configure a CDS Personal
# Access Token (PAT) once per environment, for example:
  library(ecmwfr)

wf_set_key(
  key  = "b8cda5a0-7a4d-4827-82b8-42257c851b96"
)
#
# Obtain a PAT from https://cds.climate.copernicus.eu/profile after
# accepting the ERA5-Land licence terms. Never paste the token into this
# file or commit it to Git. [web:73][web:75][web:78][web:81]
# ============================================================================
```

```{r libraries}
# Core packages required regardless of configuration. Install on first
# environment setup with:
#   install.packages(c("dplyr","tidyr","lubridate","purrr","readr",
#                      "stringr","terra","knitr","kableExtra"))
#
# Configuration-conditional packages 
#   ERA5_ACQUISITION = "cds_download"  → install.packages("ecmwfr")
#   UTCI_SOURCE      = "compute_in_script_package" → install.packages("comf")

suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(lubridate)
  library(purrr)
  library(readr)
  library(stringr)
  library(terra)
  library(knitr)
  library(kableExtra)
})

# Conditionally-loaded packages: only required for specific configurations.
# Loading them on demand keeps the script runnable in environments that
# only need a subset of capabilities.
load_if_needed <- function(pkg) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    stop("Package '", pkg, "' is required for the chosen configuration but is not installed. ",
         "Install with: install.packages('", pkg, "')")
  }
  suppressPackageStartupMessages(library(pkg, character.only = TRUE))
}

if (ERA5_ACQUISITION == "cds_download")           load_if_needed("ecmwfr")
if (UTCI_SOURCE      == "compute_in_script_package") load_if_needed("comf")
```

```{r helpers}
nice_table <- function(x, caption = NULL, full_width = FALSE,
                       wrap_col = NULL, wrap_width = "10cm") {
  fmt <- knitr::is_html_output()
  k <- kable(x, caption = caption, format.args = list(big.mark = ","),
             booktabs = TRUE, longtable = TRUE)
  if (fmt) {
    k <- kable_styling(k,
                       bootstrap_options = c("striped", "hover", "condensed"),
                       full_width = full_width, position = "left")
  } else {
    k <- kable_styling(k,
                       latex_options = c("striped", "hold_position", "repeat_header"),
                       full_width = full_width, position = "left")
  }
  if (!is.null(wrap_col)) k <- column_spec(k, wrap_col, width = wrap_width)
  k
}

# Defensive coordinate check: catches NA, infinite, and any value outside
# the catchment bounding box (a sentinel like 0/0 or a swapped lat/lon).
valid_coords <- function(lon, lat,
                         lat_min = BBOX$lat_min, lat_max = BBOX$lat_max,
                         lon_min = BBOX$lon_min, lon_max = BBOX$lon_max) {
  is.finite(lon) & is.finite(lat) &
    lon >= lon_min & lon <= lon_max &
    lat >= lat_min & lat <= lat_max
}
```

# Load the spine

```{r load-spine}
spine <- readRDS(IN_PATH)

cat(sprintf("Spine rows (= unique children): %s\n", format(nrow(spine), big.mark = ",")))
cat(sprintf("Birth-year range: %d–%d\n",
            min(spine$birth_year, na.rm = TRUE),
            max(spine$birth_year, na.rm = TRUE)))
cat(sprintf("Distinct mothers: %s\n", format(dplyr::n_distinct(spine$motherid), big.mark = ",")))
cat(sprintf("Distinct compounds: %s\n", format(dplyr::n_distinct(spine$locationid), big.mark = ",")))
```

## Coverage gate — geocodes

Pregnancies with valid household coordinates enter the household-coordinate
exposure linkage; the residual subset is held for the community-centroid
sensitivity branch (§3.10.7, item i in Chapter 3).

```{r gps-coverage}
spine <- spine |>
  mutate(coord_valid = valid_coords(gps_longitude, gps_latitude))

gps_summary <- tibble(
  metric = c("rows in spine",
             "valid household coordinates",
             "coordinates outside catchment bbox",
             "missing coordinates",
             "valid coordinate coverage (%)"),
  value  = c(nrow(spine),
             sum(spine$coord_valid, na.rm = TRUE),
             sum(!is.na(spine$gps_latitude) & !spine$coord_valid),
             sum(is.na(spine$gps_latitude) | is.na(spine$gps_longitude)),
             round(100 * mean(spine$coord_valid, na.rm = TRUE), 2))
)

nice_table(gps_summary, caption = "GPS coverage diagnostics")

# Primary exposure linkage uses only the geocoded subset.
preg <- spine |>
  filter(coord_valid,
         birth_year >= YEAR_MIN, birth_year <= YEAR_MAX) |>
  dplyr::select(child_id, motherid, locationid,
                gps_longitude, gps_latitude,
                child_dob, preg_start_assumed,
                preg_t3_start, preg_last30, preg_last7) |>
  rename(lon = gps_longitude, lat = gps_latitude) |>
  # Defensive coercion: spine.rds may store dates as Date, POSIXct, or
  # character depending on the upstream version. Coerce explicitly and
  # drop any row where either the delivery date or the reconstructed
  # pregnancy start is missing (these break seq.Date() downstream).
  mutate(
    child_dob          = as.Date(child_dob),
    preg_start_assumed = as.Date(preg_start_assumed),
    preg_t3_start      = as.Date(preg_t3_start),
    preg_last30        = as.Date(preg_last30),
    preg_last7         = as.Date(preg_last7)
  ) |>
  filter(!is.na(child_dob), !is.na(preg_start_assumed)) |>
  mutate(row_id = row_number())

cat(sprintf("\nPregnancies entering household-coordinate exposure linkage: %s\n",
            format(nrow(preg), big.mark = ",")))
```

## Unique household locations

Many pregnancies share a household coordinate. Extracting at the
location-level (rather than the pregnancy level) makes the climate-linkage
step approximately *k*-fold faster, where *k* is the average pregnancies
per location. The exposure values are joined back to the pregnancy table
via the `loc_id` index.

```{r unique-locations}
locations <- preg |>
  distinct(lon, lat) |>
  arrange(lat, lon) |>
  mutate(loc_id = row_number())

preg <- preg |>
  left_join(locations, by = c("lon", "lat"))

cat(sprintf("Unique household coordinates: %s\n",
            format(nrow(locations), big.mark = ",")))
cat(sprintf("Mean pregnancies per location: %.1f\n",
            nrow(preg) / nrow(locations)))
```

# Build the (location, day) extraction grid

For each pregnancy, the analytic window of interest is the union of:
the daily DLNM lag window (`child_dob - 30 days` to `child_dob`); the
trimester windows (`preg_start_assumed` to `child_dob`); and the
outcome-anchored short windows (already covered by the DLNM window).
The trimester windows extend further back in time than the DLNM
window. The script builds the **union** of these as the (location, day)
extraction grid.

```{r needed-days}
preg_windows <- preg |>
  mutate(
    win_start = pmin(preg_start_assumed, child_dob - days(LAG_MAX)),
    win_end   = child_dob
  ) |>
  dplyr::select(row_id, loc_id, win_start, win_end)

# Expand to one row per (pregnancy, day)
preg_days <- preg_windows |>
  rowwise() |>
  mutate(day = list(seq.Date(win_start, win_end, by = "day"))) |>
  ungroup() |>
  unnest(day) |>
  dplyr::select(row_id, loc_id, day)

# (location, day) extraction grid — what climate data must be read
extract_grid <- preg_days |>
  distinct(loc_id, day) |>
  arrange(loc_id, day)

cat(sprintf("Total (pregnancy, day) pairs needed: %s\n",
            format(nrow(preg_days), big.mark = ",")))
cat(sprintf("Unique (location, day) pairs to extract: %s\n",
            format(nrow(extract_grid), big.mark = ",")))
cat(sprintf("Compression ratio: %.2fx\n",
            nrow(preg_days) / nrow(extract_grid)))

# Save the extraction grid to disk for the climate-acquisition step.
saveRDS(extract_grid, file.path(LOG_DIR, "extract_grid.rds"))
```

# ERA5-Land input acquisition

The script supports three acquisition modes via `ERA5_ACQUISITION` in the
configuration block. Each pulls the four hourly inputs needed for UTCI:
2 m air temperature (`2m_temperature`), 2 m dewpoint temperature
(`2m_dewpoint_temperature`), 10 m u-wind and v-wind components
(`10m_u_component_of_wind`, `10m_v_component_of_wind`), and surface
downward solar radiation (`surface_solar_radiation_downwards`).

## Option 2a — Download via Copernicus CDS

```{r cds-download, eval = (ERA5_ACQUISITION == "cds_download")}
cds_request <- function(year, month) {
  first_day <- as.Date(sprintf("%04d-%02d-01", year, month))
  n_days    <- lubridate::days_in_month(first_day)

  list(
    dataset_short_name = "reanalysis-era5-land",
    variable           = c("2m_temperature",
                           "2m_dewpoint_temperature",
                           "10m_u_component_of_wind",
                           "10m_v_component_of_wind",
                           "surface_solar_radiation_downwards"),
    year               = sprintf("%04d", year),
    month              = sprintf("%02d", month),
    day                = sprintf("%02d", seq_len(n_days)),
    time               = sprintf("%02d:00", 0:23),
    area               = c(BBOX$lat_max, BBOX$lon_min,
                           BBOX$lat_min, BBOX$lon_max),
    data_format        = "netcdf",
    download_format    = "unarchived",
    target             = sprintf("era5_land_navrongo_%04d%02d.nc", year, month)
  )
}

download_one_month <- function(year, month, path, max_tries = 4) {
  req <- cds_request(year, month)

  for (attempt in seq_len(max_tries)) {
    ok <- tryCatch({
      ecmwfr::wf_request(
        request  = req,
        transfer = TRUE,
        path     = path,
        time_out = 3 * 3600,
        verbose  = FALSE
      )
      TRUE
    }, error = function(e) {
      msg <- conditionMessage(e)
      message(sprintf("Failed %04d-%02d on attempt %d/%d: %s",
                      year, month, attempt, max_tries, msg))
      FALSE
    })

    out_file <- file.path(path, sprintf("era5_land_navrongo_%04d%02d.nc", year, month))
    if (ok && file.exists(out_file)) return(TRUE)

    if (attempt < max_tries) {
      Sys.sleep(30 * attempt)
    }
  }

  FALSE
}

month_grid <- expand.grid(year = YEAR_MIN:YEAR_MAX, month = 1:12) |>
  dplyr::arrange(year, month) |>
  dplyr::mutate(
    file = sprintf("era5_land_navrongo_%04d%02d.nc", year, month),
    path = file.path(CLIMATE_DIR, file)
  ) |>
  dplyr::filter(!file.exists(path))

if (nrow(month_grid) == 0) {
  cat("All required ERA5-Land tiles already cached.\n")
} else {
  cat(sprintf("Downloading %d month-tiles from CDS...\n", nrow(month_grid)))

  failed <- vector("list", 0)

  for (i in seq_len(nrow(month_grid))) {
    yy <- month_grid$year[i]
    mm <- month_grid$month[i]

    cat(sprintf("[%d/%d] %04d-%02d\n", i, nrow(month_grid), yy, mm))

    success <- download_one_month(yy, mm, CLIMATE_DIR, max_tries = 4)

    if (!success) {
      failed[[length(failed) + 1]] <- data.frame(year = yy, month = mm)
    }

    if (i %% 6 == 0 || i == nrow(month_grid)) {
      cat(sprintf("  completed %d / %d requests\n", i, nrow(month_grid)))
    }
  }

  if (length(failed) > 0) {
    failed_df <- dplyr::bind_rows(failed)
    readr::write_csv(failed_df, file.path(LOG_DIR, "era5_failed_downloads.csv"))
    cat(sprintf("Failed months logged to %s\n",
                file.path(LOG_DIR, "era5_failed_downloads.csv")))
  } else {
    cat("All requested ERA5-Land tiles downloaded successfully.\n")
  }
}
```


# Compute daily UTCI

The Bröde et al. (2012) polynomial approximation gives UTCI as a function
of air temperature ($T_a$, °C), mean radiant temperature (MRT, °C), wind
speed at 10 m above ground ($v_a$, m/s), and relative humidity
($\text{RH}$, %). When `UTCI_SOURCE = "compute_in_script_package"` the
script uses the `comf` R package (Schweiker 2016), which implements the
Bröde polynomial as `comf::calcUTCI(ta, tr, vel, rh)`. The fully-manual
fallback (`UTCI_SOURCE = "compute_in_script_manual"`) sources a
`utci_brode.R` helper file that contains the same polynomial in
explicit form.

The mean radiant temperature is approximated following Di Napoli et al.
(2018) using only ERA5-Land 2 m air temperature and de-accumulated
downward shortwave radiation, with constants chosen for Sahelian
land-surface conditions (albedo 0.30, projected-area factor 0.97).

A subtle but consequential implementation detail: ERA5-Land's
`surface_solar_radiation_downwards` (SSRD) is accumulated in J/m² from
00:00 UTC to the forecast hour, not instantaneous W/m². To recover
hourly mean W/m² for the MRT calculation, the script de-accumulates the
SSRD field via successive differences before the unit conversion. This
step matters: treating the raw accumulated value as a per-second mean
would produce MRT estimates that are larger by approximately a factor of
the hour-of-day, with substantial downstream bias in UTCI.

## Option 1A/1C — Compute from ERA5-Land inputs

```{r compute-utci, eval = (UTCI_SOURCE != "precomputed_daily")}
# For each (year, month) tile, read the four input variables, compute
# hourly UTCI at each unique household coordinate, aggregate to daily
# summaries, and write a daily UTCI raster to UTCI_DIR.

# ----------------------------------------------------------------------------
# Helper: De-accumulate ERA5-Land hourly accumulated radiation.
# ERA5-Land surface_solar_radiation_downwards (SSRD) is accumulated from
# 00:00 UTC to the forecast hour. To recover the per-hour mean W/m² for
# each hour-ending timestamp, we take the difference between successive
# accumulated values and divide by 3600 (seconds in an hour). The 01:00
# value is the accumulation from 00:00 to 01:00 and is used directly.
# A daily reset is assumed at every 00:00 UTC timestamp.
# ----------------------------------------------------------------------------
deaccumulate_ssrd <- function(ssrd_accum_matrix, hour_of_day) {
  # ssrd_accum_matrix: locations × hours, in J/m² (accumulated since 00 UTC)
  # hour_of_day: integer 0..23 corresponding to each column of the matrix
  n_loc <- nrow(ssrd_accum_matrix)
  n_t   <- ncol(ssrd_accum_matrix)
  out   <- matrix(NA_real_, n_loc, n_t)
  for (j in seq_len(n_t)) {
    if (hour_of_day[j] == 0) {
      # 00:00 timestamp: accumulation has just reset; no preceding hour.
      out[, j] <- 0
    } else if (j == 1 || hour_of_day[j] == 1) {
      # 01:00 timestamp: accumulation from 00:00 to 01:00, use directly.
      out[, j] <- ssrd_accum_matrix[, j] / 3600
    } else {
      # All other hours: difference from previous hour, divide by 3600.
      out[, j] <- (ssrd_accum_matrix[, j] - ssrd_accum_matrix[, j - 1]) / 3600
    }
  }
  # SSRD must be non-negative; clamp small negative numerical artefacts to 0.
  out[out < 0] <- 0
  out
}

# ----------------------------------------------------------------------------
# Helper: Magnus formula for saturation vapour pressure (hPa) from T (°C).
# ----------------------------------------------------------------------------
e_sat <- function(T_C) 6.112 * exp((17.62 * T_C) / (243.12 + T_C))

# ----------------------------------------------------------------------------
# Parse the hourly time axis from an ERA5-Land SpatRaster.
# CDS NetCDFs often encode valid_time in layer names (e.g. t2m_valid_time=...)
# rather than a GDAL-readable time dimension, so terra::time() returns NA.
# ----------------------------------------------------------------------------
parse_era5_times <- function(spat_rast) {
  t <- terra::time(spat_rast)
  if (length(t) == terra::nlyr(spat_rast) && !all(is.na(t))) {
    return(t)
  }
  nm <- names(spat_rast)
  if (any(grepl("valid_time=", nm, fixed = TRUE))) {
    unix_sec <- as.numeric(sub(".*valid_time=", "", nm))
    return(as.POSIXct(unix_sec, origin = "1970-01-01", tz = "UTC"))
  }
  stop(sprintf("Could not determine time axis for %s", basename(terra::sources(spat_rast)[1])))
}

# ----------------------------------------------------------------------------
# Read one ERA5-Land variable from a NetCDF, returning a matrix of
# (locations × time) values with associated timestamps. Handles both the
# short-name and long-name variable conventions that CDS may return.
# ----------------------------------------------------------------------------
read_era5_variable <- function(nc_path, var_candidates, locations_pts) {
  # Try each candidate name in turn; the first one that opens cleanly wins.
  for (vn in var_candidates) {
    sub_r <- tryCatch(
      terra::rast(nc_path, subds = vn),
      error = function(e) NULL
    )
    if (!is.null(sub_r)) {
      pts <- terra::vect(as.matrix(locations_pts[, c("lon", "lat")]),
                         crs = "EPSG:4326")
      ex  <- terra::extract(sub_r, pts)[, -1, drop = FALSE]
      return(list(
        values    = as.matrix(ex),
        times     = parse_era5_times(sub_r),
        var_used  = vn
      ))
    }
  }
  stop(sprintf("None of the candidate variable names found in %s: %s",
               basename(nc_path), paste(var_candidates, collapse = ", ")))
}

# ----------------------------------------------------------------------------
# Main: compute daily UTCI summaries for one ERA5-Land month-tile.
# ----------------------------------------------------------------------------
compute_daily_utci <- function(nc_path, locations_pts, daily_stat) {

  # Read each variable explicitly by trying both short and long names.
  # This is more robust than parsing names() of the combined SpatRaster,
  # because the CDS layer-naming conventions have changed over time and
  # vary between the legacy and CDS-Beta API responses.
  t2m   <- read_era5_variable(nc_path,
              c("t2m", "2m_temperature"), locations_pts)
  d2m   <- read_era5_variable(nc_path,
              c("d2m", "2m_dewpoint_temperature"), locations_pts)
  u10   <- read_era5_variable(nc_path,
              c("u10", "10m_u_component_of_wind"), locations_pts)
  v10   <- read_era5_variable(nc_path,
              c("v10", "10m_v_component_of_wind"), locations_pts)
  ssrd  <- read_era5_variable(nc_path,
              c("ssrd", "surface_solar_radiation_downwards"), locations_pts)

  # Sanity check: all five variables should share the same time axis.
  stopifnot(length(t2m$times) == length(d2m$times),
            length(t2m$times) == length(u10$times),
            length(t2m$times) == length(v10$times),
            length(t2m$times) == length(ssrd$times))

  # Unit conversions
  t2m_C   <- t2m$values - 273.15        # K → °C
  d2m_C   <- d2m$values - 273.15        # K → °C
  wind_ms <- sqrt(u10$values^2 + v10$values^2)

  # De-accumulate SSRD: ERA5-Land returns accumulated J/m² since 00 UTC.
  # Recover per-hour mean W/m² via finite differences before MRT.
  hour_of_day <- lubridate::hour(t2m$times)
  ssrd_Wm2    <- deaccumulate_ssrd(ssrd$values, hour_of_day)

  # Relative humidity from dewpoint, in percent. comf::calcUTCI() expects RH%.
  # RH = 100 × e_sat(T_d) / e_sat(T_a)
  rh_pct <- 100 * e_sat(d2m_C) / e_sat(t2m_C)
  rh_pct[rh_pct > 100] <- 100  # clamp numerical artefacts

  # Mean radiant temperature (Di Napoli et al. 2018 simplified form for
  # gridded reanalysis inputs). MRT in °C from absorbed shortwave plus
  # ambient longwave, using land-surface constants for Sahelian conditions:
  #   albedo ≈ 0.30, projected-area factor ε_p ≈ 0.97, σ = 5.67e-8 W/m²/K⁴.
  sigma  <- 5.67e-8
  albedo <- 0.30
  eps_p  <- 0.97
  mrt_K  <- ((ssrd_Wm2 * (1 - albedo) / (sigma * eps_p)) +
             (t2m_C + 273.15)^4)^0.25
  mrt_C  <- mrt_K - 273.15

  # Hourly UTCI computation.
  n_loc <- nrow(t2m_C)
  n_t   <- ncol(t2m_C)
  utci_hourly <- matrix(NA_real_, n_loc, n_t)

  if (UTCI_SOURCE == "compute_in_script_package") {
    # comf::calcUTCI(ta, tr, vel, rh) — air temperature (°C), mean radiant
    # temperature (°C), wind speed at 10 m (m/s), relative humidity (%).
    for (j in seq_len(n_t)) {
      utci_hourly[, j] <- comf::calcUTCI(
        ta  = t2m_C[, j],
        tr  = mrt_C[, j],
        vel = wind_ms[, j],
        rh  = rh_pct[, j]
      )
    }
  } else {
    # Manual Bröde polynomial fallback (sourced from a separate helper).
    if (file.exists("utci_brode.R")) {
      source("utci_brode.R")
      utci_hourly <- utci_brode(t2m_C, mrt_C, wind_ms, rh_pct)
    } else {
      stop("UTCI_SOURCE = 'compute_in_script_manual' but utci_brode.R is not present.")
    }
  }

  # Aggregate to daily summary at each location.
  # PRESERVE the actual loc_id from locations_pts rather than re-indexing.
  hour_dates <- as.Date(t2m$times)
  unique_days <- sort(unique(hour_dates))

  daily <- map_dfr(unique_days, function(d) {
    cols_today <- which(hour_dates == d)
    if (length(cols_today) == 0) return(NULL)
    if (daily_stat == "max") {
      vals <- apply(utci_hourly[, cols_today, drop = FALSE], 1,
                    max, na.rm = TRUE)
    } else if (daily_stat == "mean") {
      vals <- rowMeans(utci_hourly[, cols_today, drop = FALSE], na.rm = TRUE)
    } else {
      stop("Unknown daily_stat: ", daily_stat)
    }
    # Replace -Inf (which arises when max() is applied to all-NA rows) with NA.
    vals[!is.finite(vals)] <- NA_real_
    tibble(loc_id = locations_pts$loc_id, day = d, utci = vals)
  })

  daily
}

# Iterate over month-tiles, computing daily UTCI and accumulating.
month_files <- list.files(CLIMATE_DIR,
                          pattern = "^era5_land_navrongo_\\d{6}\\.nc$",
                          full.names = TRUE)

daily_utci_list <- vector("list", length(month_files))
start_time <- Sys.time()

for (k in seq_along(month_files)) {
  daily_utci_list[[k]] <- compute_daily_utci(
    nc_path        = month_files[k],
    locations_pts  = locations,
    daily_stat     = DAILY_STAT_PRIMARY
  )
  if (k %% 12 == 0 || k == length(month_files)) {
    elapsed <- difftime(Sys.time(), start_time, units = "mins")
    cat(sprintf("  Tile %d / %d — %.1f min elapsed\n",
                k, length(month_files), as.numeric(elapsed)))
  }
}

daily_utci <- bind_rows(daily_utci_list)

# Write a single consolidated daily UTCI table for downstream use.
saveRDS(daily_utci, file.path(UTCI_DIR, "daily_utci_consolidated.rds"))
cat(sprintf("\nDaily UTCI table written: %s rows\n",
            format(nrow(daily_utci), big.mark = ",")))
```


# Assemble pregnancy-level exposure matrix

For each pregnancy in the analytic cohort, we now construct the columns
that the downstream DLNM, case-crossover, and trimester-regression scripts
consume.

```{r assemble-exposure}
# Join daily UTCI values to the (pregnancy, day) table.
preg_daily <- preg_days |>
  left_join(daily_utci, by = c("loc_id", "day"))

# Diagnose any (pregnancy, day) gaps; these become NA in the lag matrix.
n_total  <- nrow(preg_daily)
n_filled <- sum(!is.na(preg_daily$utci))
cat(sprintf("Pregnancy-day cells filled: %s of %s (%.2f%%)\n",
            format(n_filled, big.mark = ","),
            format(n_total,  big.mark = ","),
            100 * n_filled / n_total))

# ----------------------------------------------------------------------------
# DLNM daily lag matrix: one row per pregnancy, columns lag_0 ... lag_LAG_MAX.
# ----------------------------------------------------------------------------
lag_matrix <- preg_daily |>
  inner_join(preg |> dplyr::select(row_id, child_dob), by = "row_id") |>
  mutate(lag = as.integer(child_dob - day)) |>
  filter(lag >= LAG_MIN, lag <= LAG_MAX) |>
  pivot_wider(id_cols     = row_id,
              names_from  = lag,
              values_from = utci,
              names_prefix = "lag_")

# Ensure column order is lag_0, lag_1, ..., lag_30
lag_cols <- paste0("lag_", LAG_MIN:LAG_MAX)
missing_lag_cols <- setdiff(lag_cols, names(lag_matrix))
for (m in missing_lag_cols) lag_matrix[[m]] <- NA_real_
lag_matrix <- lag_matrix[, c("row_id", lag_cols)]

# ----------------------------------------------------------------------------
# Trimester and full-pregnancy means.
# T1: days 0–91 post-conception   = (preg_start_assumed + 0)  ... (preg_start_assumed + 91)
# T2: days 92–182                 = (preg_start_assumed + 92) ... (preg_start_assumed + 182)
# T3: days 183–delivery           = (preg_start_assumed + 183)... child_dob
# Full pregnancy: preg_start_assumed ... child_dob
# ----------------------------------------------------------------------------
preg_with_dates <- preg_daily |>
  inner_join(preg |> dplyr::select(row_id, child_dob, preg_start_assumed),
             by = "row_id") |>
  mutate(days_from_conception = as.integer(day - preg_start_assumed))

trimester_means <- preg_with_dates |>
  mutate(trimester = case_when(
    days_from_conception >= 0   & days_from_conception <= TRIMESTER1_END ~ "t1",
    days_from_conception > TRIMESTER1_END & days_from_conception <= TRIMESTER2_END ~ "t2",
    days_from_conception > TRIMESTER2_END & day <= child_dob ~ "t3",
    TRUE ~ NA_character_
  )) |>
  filter(!is.na(trimester)) |>
  group_by(row_id, trimester) |>
  summarise(mean_utci = if (all(is.na(utci))) NA_real_ else mean(utci, na.rm = TRUE),
            .groups = "drop") |>
  pivot_wider(names_from = trimester, values_from = mean_utci,
              names_prefix = "")

preg_mean <- preg_with_dates |>
  filter(day >= preg_start_assumed, day <= child_dob) |>
  group_by(row_id) |>
  summarise(preg_mean_utci = if (all(is.na(utci))) NA_real_ else mean(utci, na.rm = TRUE),
            .groups = "drop")

# ----------------------------------------------------------------------------
# Outcome-anchored short-window means (last 7, 14, 30 days).
# ----------------------------------------------------------------------------
short_window_mean <- function(df, n_days, varname) {
  df |>
    mutate(lag = as.integer(child_dob - day)) |>
    filter(lag >= 0, lag < n_days) |>
    group_by(row_id) |>
    summarise(!!varname := if (all(is.na(utci))) NA_real_ else mean(utci, na.rm = TRUE),
              .groups = "drop")
}

short_windows <- reduce(
  map(OUTCOME_WINDOWS, function(n) {
    short_window_mean(preg_with_dates, n, paste0("last_", n, "d_mean_utci"))
  }),
  function(a, b) full_join(a, b, by = "row_id")
)

# ----------------------------------------------------------------------------
# Coverage diagnostic per pregnancy.
# ----------------------------------------------------------------------------
coverage <- lag_matrix |>
  mutate(lag_cells_filled = rowSums(!is.na(across(starts_with("lag_")))),
         lag_coverage_pct = round(100 * lag_cells_filled / (LAG_MAX - LAG_MIN + 1), 1)) |>
  dplyr::select(row_id, lag_cells_filled, lag_coverage_pct)

# ----------------------------------------------------------------------------
# Assemble the final exposure table.
# ----------------------------------------------------------------------------
exposure <- preg |>
  dplyr::select(row_id, child_id, motherid, locationid, lon, lat,
                child_dob, preg_start_assumed) |>
  left_join(lag_matrix,        by = "row_id") |>
  left_join(trimester_means |>
              rename(t1_mean_utci = t1, t2_mean_utci = t2, t3_mean_utci = t3),
            by = "row_id") |>
  left_join(preg_mean,         by = "row_id") |>
  left_join(short_windows,     by = "row_id") |>
  left_join(coverage,          by = "row_id")
```

# Diagnostics

```{r diag-tables}
exposure_vars <- c("t1_mean_utci", "t2_mean_utci", "t3_mean_utci",
                   "preg_mean_utci",
                   paste0("last_", OUTCOME_WINDOWS, "d_mean_utci"))

# Coverage summary
coverage_summary <- tibble(
  variable = exposure_vars,
  pct_non_missing = sapply(exposure_vars, function(v)
    round(100 * mean(!is.na(exposure[[v]])), 1)),
  median_utci = sapply(exposure_vars, function(v)
    round(median(exposure[[v]], na.rm = TRUE), 1)),
  p10 = sapply(exposure_vars, function(v)
    round(quantile(exposure[[v]], 0.10, na.rm = TRUE), 1)),
  p90 = sapply(exposure_vars, function(v)
    round(quantile(exposure[[v]], 0.90, na.rm = TRUE), 1))
)

nice_table(coverage_summary,
           caption = "Per-pregnancy exposure summary — coverage and distribution of derived window means")

# DLNM lag-matrix coverage
lag_coverage_summary <- tibble(
  metric = c("pregnancies with 100% lag coverage",
             "pregnancies with ≥95% lag coverage",
             "pregnancies with ≥80% lag coverage",
             "median lag coverage (%)"),
  value  = c(sum(exposure$lag_coverage_pct == 100),
             sum(exposure$lag_coverage_pct >= 95),
             sum(exposure$lag_coverage_pct >= 80),
             round(median(exposure$lag_coverage_pct), 1))
)
nice_table(lag_coverage_summary,
           caption = "DLNM lag-matrix coverage (lag_0 through lag_30)")
```

# Persist outputs

```{r save}
saveRDS(exposure, file.path(OUT_DIR, "exposure.rds"))
write_csv(exposure |> dplyr::select(-starts_with("lag_")),
          file.path(OUT_DIR, "exposure_summaries.csv"))

cat(sprintf("Files written:\n  %s\n  %s\n",
            file.path(OUT_DIR, "exposure.rds"),
            file.path(OUT_DIR, "exposure_summaries.csv")))
```

# Expected diagnostic counts

The script has been developed against the development sample produced by
`01_build_spine.Rmd`. Sample-stage and full-data expected counts are
documented below; the full-data figures will be reported once the script
is executed against the full extract by NHRC analysts.

```{r expected, echo = FALSE}
tibble(
  metric = c("pregnancies entering exposure linkage (full extract)",
             "unique household coordinates (full extract)",
             "(location, day) pairs to extract",
             "ERA5-Land month-tiles required",
             "expected (pregnancy, day) cells in lag matrix",
             "expected (pregnancy, day) cells in pregnancy window"),
  expected_value = c("approximately 94,649 (97.5% of 97,045)",
                     "approximately 4,000–6,000",
                     "approximately 23,500,000",
                     "312 (1995–2020 × 12 months)",
                     paste0("approximately ", format(94649 * 31, big.mark = ",")),
                     paste0("approximately ", format(94649 * 280, big.mark = ",")))
) |>
  nice_table(caption = "Expected diagnostic counts at full-data execution",
             wrap_col = 1, wrap_width = "9cm")
```

# Session information

```{r session-info}
sessionInfo()
```

# Next steps

With `exposure.rds` in hand, the immediate items are, in order:

1. **Validate the exposure construction** against substantive checks: the
   seasonal pattern of daily UTCI within the catchment (peaks in
   hot-dry season, troughs in dry-cool harmattan), the spatial pattern
   across the catchment (which should be modest given the small
   geographic extent), and the year-on-year trend (which should track
   the regional warming signal). These checks belong in
   `03_phase1_descriptive.Rmd` rather than this script.

2. **Construct the DLNM cross-basis** in `05_phase3_dlnm_perinatal.Rmd`.
   The cross-basis takes the daily lag matrix from `exposure.rds` and
   produces a set of model-matrix columns via `dlnm::crossbasis()`. The
   cross-basis is the substantive output that enters the Bayesian DLNM in
   Phase 3.

3. **Run the community-centroid sensitivity branch** in
   `09_phase7_sensitivity.Rmd`. For the 2.5% of pregnancies without
   household coordinates, the climate linkage uses a community-centroid
   coordinate derived from `locationid`-level aggregation of the geocoded
   subset. The structure of `02_link_era5_utci.Rmd` is reusable for this
   branch with a single change at the GPS-coverage gate.
