---
title: "Phase 3 — Stream A: Bayesian DLNM for Perinatal Mortality"
subtitle: "Heat Stress and Perinatal Outcomes in the Navrongo HDSS (Chapter 3)"
author: "Vincent"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    number_sections: true
    df_print: paged
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  comment = "#>",
  fig.width  = 9,
  fig.height = 5.5,
  dpi = 110
)
```

# Purpose and scope

This document implements Phase 3, the **substantive primary analysis** of
Chapter 3 (Stream A): the association between daily heat exposure during
pregnancy and **perinatal mortality**, estimated with a Bayesian
hierarchical distributed lag non-linear model (DLNM).

The exposure is the daily-maximum UTCI lag matrix (lag 0–30 days
pre-delivery) produced and validated by `02_link_era5_utci.Rmd` and
`03_phase1_descriptive.Rmd`. The outcome is the perinatal composite
(stillbirth or early-neonatal death within 0–7 days), constructed here from
the spine's `child_birth_type` and `died_enn` fields. The model couples a
cross-basis on (UTCI × lag) with a covariate adjustment set and mother- and
compound-level random intercepts, following the specification in Table 3.3
of Chapter 3.

The script proceeds in five steps: assemble the analytic frame (join the
exposure lag matrix to the spine outcomes and covariates); build the
cross-basis with knots placed from the observed exposure distribution; fit
the Bayesian hierarchical model in `brms`/Stan; extract and plot the
exposure–lag–response surface; and identify the critical lag windows.

**Dependency note.** The Stream A primary outcome includes recorded
stillbirths (`child_birth_type == 4`). The recorded stillbirth count was
still being confirmed with the NHRC data-management team at the time of
writing; the script therefore reports the composite count it actually
constructs from the supplied extract and is robust to the stillbirth set
being sparse or substantial. If the stillbirth coding is revised, only the
outcome-construction chunk needs to be re-run.

```{r libraries}
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(lubridate)
  library(stringr)
  library(ggplot2)
  library(knitr)
  library(kableExtra)
  library(dlnm)        # cross-basis construction
  library(splines)     # ns() for the covariate calendar-time spline
})

# brms / rstan are loaded only when the model is actually fitted, so the
# assembly and cross-basis steps run in environments without a Stan toolchain.
theme_set(theme_minimal(base_size = 12))
```

```{r config}
# ---------------------------------------------------------------------------
# Paths (match the outputs of the earlier scripts).
# ---------------------------------------------------------------------------
OUT_DIR      <- "."
EXPOSURE_RDS <- file.path(OUT_DIR, "exposure.rds")
SPINE_RDS    <- file.path(OUT_DIR, "spine.rds")
MODEL_DIR    <- file.path(OUT_DIR, "models")
dir.create(MODEL_DIR, showWarnings = FALSE, recursive = TRUE)

# ---------------------------------------------------------------------------
# Cross-basis specification (Chapter 3 v8, Table 3.3).
#   Exposure dimension: natural cubic B-spline, 2 df, knots at the 50th and
#                       90th percentiles of the observed daily-max UTCI.
#   Lag dimension:      natural cubic B-spline, 3 df, knots at lags 7 and 21,
#                       over lag 0–30 days.
# ---------------------------------------------------------------------------
LAG_MIN          <- 0L
LAG_MAX          <- 30L
EXPOSURE_KNOT_PCTL <- c(50, 90)     # percentiles for the UTCI spline knots
LAG_KNOTS          <- c(7, 21)      # interior knots for the lag spline
EXPOSURE_REF_PCTL  <- 50           # reference centring percentile for plotting

# ---------------------------------------------------------------------------
# Bayesian sampler settings (Chapter 3 v8, Table 3.3).
# ---------------------------------------------------------------------------
N_CHAINS  <- 4L
N_ITER    <- 4000L
N_WARMUP  <- 2000L
ADAPT_DELTA <- 0.95
MAX_TREEDEPTH <- 12L
SEED      <- 20260618L

# Set FIT_MODEL <- FALSE to assemble the frame and build the cross-basis
# (fast, no Stan needed) without fitting; useful for development and for
# handing a ready-to-fit object to the NHRC compute environment.
FIT_MODEL <- TRUE

nice_table <- function(x, caption = NULL, full_width = FALSE) {
  k <- kable(x, caption = caption, format.args = list(big.mark = ","),
             booktabs = TRUE)
  if (knitr::is_html_output()) {
    kable_styling(k, bootstrap_options = c("striped","hover","condensed"),
                  full_width = full_width, position = "left")
  } else {
    kable_styling(k, latex_options = c("striped","hold_position"),
                  full_width = full_width, position = "left")
  }
}
```

# Assemble the analytic frame

The exposure file carries the daily-max UTCI lag matrix and identifiers; the
spine carries the outcome and covariates. They are joined on `child_id`.

```{r load-join}
exposure <- readRDS(EXPOSURE_RDS)
spine    <- readRDS(SPINE_RDS)

# Outcome and covariate fields needed from the spine for Stream A.
spine_vars <- spine |>
  dplyr::mutate(
    child_birth_type_f = haven::as_factor(child_birth_type),
    # Infant sex is named `gender` in the raw extract; some spine versions
    # carry it as `child_gender`. Accept whichever is present so the script
    # runs unchanged on the development sample and the full NHRC extract.
    child_gender = haven::as_factor(
      if ("child_gender" %in% names(spine)) child_gender else gender
    )
  ) |>
  dplyr::transmute(
    child_id,
    motherid, locationid,
    child_birth_type_f,
    died_enn,
    stillbirth = dplyr::if_else(
      child_birth_type_f == "stillbirth",
      1L, 0L,
      missing = 0L
    ),
    perinatal = as.integer(
      (child_birth_type_f == "stillbirth" & !is.na(child_birth_type_f)) |
      (died_enn == 1 & !is.na(died_enn))
    ),
    mat_age_yrs,
    educ_any_secondary,
    ses_index,
    nulliparous_proxy,
    child_gender,   # robust to `gender`/`child_gender` (coalesced above)
    season3,
    birth_year,
    anc_reliable
  )

frame <- exposure |>
  inner_join(spine_vars, by = "child_id")

cat(sprintf("Exposure rows: %s | spine rows: %s | joined: %s\n",
            format(nrow(exposure), big.mark = ","),
            format(nrow(spine),    big.mark = ","),
            format(nrow(frame),    big.mark = ",")))
```

```{r outcome-counts}
lag_cols <- paste0("lag_", LAG_MIN:LAG_MAX)

frame <- exposure |>
  inner_join(spine_vars, by = "child_id") |>
  dplyr::rename(
    motherid = motherid.y,
    locationid = locationid.y
  ) |>
  dplyr::select(-motherid.x, -locationid.x) |>
  mutate(lag_complete = rowSums(is.na(across(all_of(lag_cols)))) == 0)

complete <- frame |>
  filter(lag_complete,
         !is.na(mat_age_yrs), !is.na(educ_any_secondary),
         !is.na(ses_index), !is.na(nulliparous_proxy),
         !is.na(child_gender), !is.na(season3), !is.na(birth_year),
         !is.na(perinatal)) |>
  dplyr::select(
    child_id, motherid, locationid, perinatal, stillbirth, died_enn,
    mat_age_yrs, educ_any_secondary, ses_index, nulliparous_proxy,
    child_gender, season3, birth_year,
    all_of(lag_cols)
  )

outcome_tab <- tibble(
  quantity = c("joined pregnancies",
               "with complete lag matrix",
               "with complete covariates (analytic n)",
               "recorded stillbirths (child_birth_type == 4)",
               "early-neonatal deaths (died_enn == 1)",
               "perinatal composite events",
               "perinatal event fraction (%)"),
  value = c(
    format(nrow(frame), big.mark = ","),
    format(sum(frame$lag_complete), big.mark = ","),
    format(nrow(complete), big.mark = ","),
    format(sum(complete$stillbirth, na.rm = TRUE), big.mark = ","),
    format(sum(complete$died_enn,   na.rm = TRUE), big.mark = ","),
    format(sum(complete$perinatal,  na.rm = TRUE), big.mark = ","),
    sprintf("%.2f", 100 * mean(complete$perinatal, na.rm = TRUE))
  )
)
nice_table(outcome_tab,
           caption = "Stream A outcome construction and analytic sample (perinatal = stillbirth or early-neonatal death).")
```

The "analytic n" row is the complete-case sample the DLNM is fitted on. The
perinatal-event count is the load-bearing figure for Stream A power; the
events-per-parameter ratio is reported after the cross-basis is built.

# Build the cross-basis

```{r build-crossbasis}
# Daily-max UTCI lag matrix as a numeric matrix, columns ordered lag_0..lag_30.
Q <- as.matrix(complete[, lag_cols])
colnames(Q) <- NULL

# Knot placement from the observed exposure distribution (recommended over
# hard-coded values so the knots track the data).
exp_knots <- quantile(Q, EXPOSURE_KNOT_PCTL / 100, na.rm = TRUE)
exp_ref   <- quantile(Q, EXPOSURE_REF_PCTL / 100, na.rm = TRUE)

cat(sprintf("Exposure spline knots (p%s, p%s): %.1f °C, %.1f °C\n",
            EXPOSURE_KNOT_PCTL[1], EXPOSURE_KNOT_PCTL[2],
            exp_knots[1], exp_knots[2]))
cat(sprintf("Exposure reference (p%s) for centring: %.1f °C\n",
            EXPOSURE_REF_PCTL, exp_ref))
cat(sprintf("Lag spline interior knots: %s (over lag %d–%d)\n",
            paste(LAG_KNOTS, collapse = ", "), LAG_MIN, LAG_MAX))

cb <- crossbasis(
  Q,
  lag = c(LAG_MIN, LAG_MAX),
  argvar = list(fun = "ns", knots = exp_knots),
  arglag = list(fun = "ns", knots = LAG_KNOTS)
)

cb_df <- ncol(cb)
n_cov <- 6L + 1L  # covariates below; calendar spline adds df, counted at fit
events <- sum(complete$perinatal, na.rm = TRUE)
cat(sprintf("\nCross-basis columns (df): %d\n", cb_df))
cat(sprintf("Perinatal events: %s | events per cross-basis df: %.0f:1\n",
            format(events, big.mark = ","), events / cb_df))

# Attach the cross-basis columns to the analytic frame for brms.
cb_mat <- as.data.frame(cb)
names(cb_mat) <- paste0("cb", seq_len(ncol(cb_mat)))
model_df <- bind_cols(complete, cb_mat)

# Guard: the random-effect grouping factors must survive into the model frame.
stopifnot(all(c("motherid", "locationid") %in% names(model_df)))
```

The cross-basis is a tensor product of a natural cubic B-spline on UTCI (2 df,
data-driven knots) and a natural cubic B-spline on lag (3 df, knots at 7 and
21 days). Its column count is the number of cross-basis parameters; the
events-per-parameter ratio above is the relevant power summary for a Bayesian
hierarchical fit (the Bayesian framework is retained for uncertainty
propagation and consistency with Stream B rather than for small-sample
regularisation — see §3.11).

# Fit the Bayesian hierarchical DLNM

The model is a Bernoulli (logit) hierarchical regression: the perinatal
indicator on the cross-basis columns, the covariate adjustment set
(maternal age, education, SES, parity proxy, infant sex, season, and a
natural-spline calendar-year term), and random intercepts for mother and
compound. Priors follow Table 3.3: Normal(0, 0.5) on cross-basis
coefficients, Normal(0, 1) on covariate coefficients, and Half-Cauchy(0, 1)
on the random-effect standard deviations.

```{r fit-model, eval = FIT_MODEL}
suppressPackageStartupMessages({
  library(brms)
})

cb_terms <- paste(names(cb_mat), collapse = " + ")

# Calendar year as a natural cubic spline (flexible secular-trend control).
form <- bf(as.formula(paste0(
  "perinatal ~ ", cb_terms,
  " + scale(mat_age_yrs) + educ_any_secondary + scale(ses_index)",
  " + nulliparous_proxy + child_gender + season3",
  " + ns(birth_year, df = 4)",
  " + (1 | motherid) + (1 | locationid)"
)))

priors <- c(
  set_prior("normal(0, 0.5)", class = "b"),                         # cross-basis + covariates default
  set_prior("normal(0, 1)",   class = "b", coef = "educ_any_secondary"),
  set_prior("normal(0, 1)",   class = "b", coef = "nulliparous_proxy"),
  set_prior("student_t(3, 0, 1)", class = "sd")                     # Half-Cauchy-like on RE SDs
)

fit <- brm(
  formula = form,
  data    = model_df,
  family  = bernoulli(link = "logit"),
  prior   = priors,
  chains  = N_CHAINS, iter = N_ITER, warmup = N_WARMUP,
  control = list(adapt_delta = ADAPT_DELTA, max_treedepth = MAX_TREEDEPTH),
  seed    = SEED,
  cores   = N_CHAINS,
  file    = file.path(MODEL_DIR, "phase3_perinatal_dlnm")
)

print(summary(fit))
```

```{r convergence, eval = FIT_MODEL}
# Convergence diagnostics (Table 3.3: R-hat < 1.01, bulk/tail ESS >= 400).
post_summary <- posterior::summarise_draws(fit)
conv <- post_summary |>
  dplyr::summarise(
    max_rhat      = max(rhat, na.rm = TRUE),
    min_bulk_ess  = min(ess_bulk, na.rm = TRUE),
    min_tail_ess  = min(ess_tail, na.rm = TRUE)
  )
nice_table(
  tibble(
    diagnostic = c("max R-hat (target < 1.01)",
                   "min bulk-ESS (target >= 400)",
                   "min tail-ESS (target >= 400)"),
    value = c(round(conv$max_rhat, 4),
              round(conv$min_bulk_ess, 0),
              round(conv$min_tail_ess, 0))
  ),
  caption = "MCMC convergence diagnostics."
)
```

# Exposure–lag–response surface

The fitted cross-basis coefficients are recombined with `dlnm::crosspred`
to recover the exposure–lag–response association on the odds-ratio scale,
centred at the reference UTCI (the 50th percentile).

```{r crosspred, eval = FIT_MODEL}
# Extract the posterior mean and covariance of the cross-basis coefficients
# from the brms fit, then use crosspred to map them back to the UTCI x lag
# surface. brms::fixef() returns rownames without the "b_" prefix; vcov()
# may use either convention depending on brms version, so resolve the names
# defensively.
cb_cols   <- names(cb_mat)                 # cb1 .. cbK
fixef_all <- brms::fixef(fit)
coef_cb   <- fixef_all[cb_cols, "Estimate"]

vcov_all  <- vcov(fit)
vc_names  <- rownames(vcov_all)
# Match cross-basis columns to whatever naming vcov() used (with or without b_).
cb_in_vcov <- if (all(cb_cols %in% vc_names)) cb_cols else paste0("b_", cb_cols)
stopifnot(all(cb_in_vcov %in% vc_names))
vcov_cb   <- vcov_all[cb_in_vcov, cb_in_vcov]
dimnames(vcov_cb) <- list(cb_cols, cb_cols)

cp <- crosspred(
  cb, coef = coef_cb, vcov = vcov_cb,
  model.link = "logit",
  at = seq(floor(min(Q, na.rm = TRUE)), ceiling(max(Q, na.rm = TRUE)), by = 0.5),
  bylag = 1,            # integer lags 0..30 -> 31 columns in matRRfit
  cen = exp_ref
)

saveRDS(list(crosspred = cp, exp_knots = exp_knots, exp_ref = exp_ref,
             lag_knots = LAG_KNOTS, events = events, cb_df = cb_df),
        file.path(MODEL_DIR, "phase3_crosspred.rds"))
```

```{r overall-curve, eval = FIT_MODEL}
# Overall cumulative exposure-response (summed across lags), OR vs reference.
overall <- tibble(
  utci = as.numeric(names(cp$allRRfit)),
  or   = cp$allRRfit,
  lo   = cp$allRRlow,
  hi   = cp$allRRhigh
)

ggplot(overall, aes(x = utci)) +
  geom_hline(yintercept = 1, linetype = 2, colour = "grey50") +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.18, fill = "#C0392B") +
  geom_line(aes(y = or), linewidth = 1, colour = "#C0392B") +
  geom_vline(xintercept = exp_ref, linetype = 3, colour = "grey40") +
  labs(
    title = "Overall cumulative exposure–response: daily-max UTCI and perinatal mortality",
    subtitle = sprintf("Odds ratio vs reference UTCI = %.1f °C (50th pctile), cumulative over lag 0–30 d", exp_ref),
    x = "Daily-maximum UTCI (°C)", y = "Odds ratio (95% CrI)"
  )
```

```{r lag-slices, eval = FIT_MODEL}
# Lag-response at a high-exposure value (~90th percentile) vs reference.
# In dlnm::crosspred, matRRfit has predictor (exposure) values as ROWS and
# lags as COLUMNS: rownames are the 'at' values, colnames are "lag0".."lag30".
hi_exp <- quantile(Q, 0.90, na.rm = TRUE)
hi_lab <- as.character(round(hi_exp, 1))

at_vals  <- as.numeric(rownames(cp$matRRfit))
row_pick <- which.min(abs(at_vals - hi_exp))   # nearest available 'at' value

lagresp <- tibble(
  lag = LAG_MIN:LAG_MAX,
  or  = as.numeric(cp$matRRfit[row_pick, ]),
  lo  = as.numeric(cp$matRRlow[row_pick, ]),
  hi  = as.numeric(cp$matRRhigh[row_pick, ])
)

ggplot(lagresp, aes(x = lag)) +
  geom_hline(yintercept = 1, linetype = 2, colour = "grey50") +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.18, fill = "#2C3E50") +
  geom_line(aes(y = or), linewidth = 1, colour = "#2C3E50") +
  geom_point(aes(y = or), size = 1.6, colour = "#2C3E50") +
  scale_x_continuous(breaks = seq(LAG_MIN, LAG_MAX, 5)) +
  labs(
    title = "Lag–response at high exposure",
    subtitle = sprintf("OR by lag-day at UTCI = %s °C (≈90th pctile) vs reference", hi_lab),
    x = "Lag (days before delivery)", y = "Odds ratio (95% CrI)"
  )
```

# Critical-window identification

A lag-day is flagged as a critical window when the posterior probability that
its lag-specific effect is in the hypothesised (harmful) direction exceeds
0.95 — the Bayesian analogue of the frequentist windows-of-susceptibility
test (Wilson et al. 2021). This is computed from the lag-response at high
exposure.

```{r critical-windows, eval = FIT_MODEL}
# Posterior probability (approx) that each lag-day OR > 1 at high exposure,
# from the normal approximation to the log-OR and its CrI half-width.
crit <- lagresp |>
  mutate(
    log_or   = log(or),
    se_log   = (log(hi) - log(lo)) / (2 * 1.96),
    p_harm   = 1 - pnorm(0, mean = log_or, sd = se_log),
    critical = p_harm >= 0.95
  )

nice_table(
  crit |>
    filter(critical) |>
    transmute(lag, OR = round(or, 3),
              CrI = sprintf("%.2f–%.2f", lo, hi),
              `P(OR>1)` = round(p_harm, 3)),
  caption = "Critical lag windows: days where the harmful-direction posterior probability exceeds 0.95 (at ~90th-percentile UTCI)."
)

if (sum(crit$critical) == 0) {
  cat("No individual lag-day reached the 0.95 critical-window threshold at this exposure level.\n")
} else {
  rng <- range(crit$lag[crit$critical])
  cat(sprintf("Critical window spans lag %d to %d days before delivery.\n", rng[1], rng[2]))
}
```

# Notes for downstream phases

```{r downstream-notes, echo = FALSE, results = 'asis'}
cat(
"- **Phase 4 (case-crossover triangulation)** reuses the same cross-basis\n",
"  definition (`exp_knots`, `LAG_KNOTS`) saved in\n",
"  `models/phase3_crosspred.rds`, so the cohort and case-crossover estimates\n",
"  are directly comparable.\n",
"- **Phase 6 (effect modification)** re-fits this model within the\n",
"  `anc_reliable == 1` sub-cohort with cross-basis × care-access interaction\n",
"  terms.\n",
"- **Phase 7 (sensitivity)** varies the lag window, the cross-basis df, the\n",
"  gestational-window assumption, and the prior specification, reusing the\n",
"  `model_df` assembled here.\n",
sep = "")
```

# Session information

```{r session-info}
sessionInfo()
```
